Sex Differences in Discrimination Behavior and Orbitofrontal Engagement During Context-Gated Reward Prediction

Version 1

Abstract

Animals, including humans, rely on contextual information to interpret ambiguous stimuli. Impaired context processing is a hallmark of several neuropsychiatric disorders, including schizophrenia, autism, PTSD, and addiction. While sex differences in the prevalence and manifestations of these disorders are well established, potential sex differences in context processing remain uncertain. Here we examined sex differences in the contextual control over cue-evoked reward seeking and its neural correlates, in rats. Male and female rats were trained in a bidirectional occasion-setting preparation in which the validity of two reward-predictive cues was informed by the presence, or absence, of a background contextual feature (A:X+ / X-/ A:Y-/ Y+). Females were significantly slower to acquire contextual control over cue-evoked reward seeking. However, once established, the contextual control over behavior was more robust in female rats; it showed less within-session variability (less influence of prior reward) and greater resistance to acute stress. This superior contextual control achieved by females was accompanied by an increased activation of the orbitofrontal cortex compared to males. Critically, these behavioral and neural sex differences were specific to the contextual modulation process and not observed in simple, context-independent, reward prediction tasks. These results indicate a sex-biased trade-off between the speed of acquisition and the robustness of performance in the contextual modulation of cued reward seeking. The different distribution of sexes along the Fast learning ↔ Steady performance continuum might reflect different levels of engagement of the orbitofrontal cortex, and might have implications for our understanding of sex differences in psychiatric disorders.

Introduction

Pavlovian learning is often conceived as the formation of a binary association between two events, where event #1 (the stimulus) evokes the anticipation of event #2 (the outcome). Under this strict definition, the background circumstances (the context) in which those events occur play little role. However, this simple form of learning is often augmented by higher level cognitive systems that allow animals (including humans) to learn and retrieve context-specific associations (e.g., the word “shot” might evoke different representations in a bar vs. a vaccine clinic) [1–3].

Deficits in contextual modulation are observed in several neuropsychiatric disorders including schizophrenia, autism, post-traumatic stress disorders (PTSD), and substance use disorders (SUD) [4–11], and may contribute to the intrusive thoughts and maladaptive (situation-inappropriate) behaviors that characterize these disorders. Importantly, the prevalence and symptomatology of these disorders vary between sexes. For example, the incidence of schizophrenia and autism is higher in men [12] while women are more at risk of developing SUD or PTSD following initial drug use or trauma [13, 14]. This disparity might be due, in part, to subtle sex differences in executive strategies [15–17]. Stress, a contributing factor in all aforementioned disorders, can further exacerbate these sex differences, resulting in sex-specific cognitive vulnerabilities [17–19]. Sex differences in executive functions and sensitivity to stress have been reproduced in animal models using behavioral assays of attention, impulsivity, and working memory [16, 20, 21]. Whether these sex differences extend to the contextual modulation of associative predictions remains unknown.

In animal models, the contextual modulation of associative predictions can be studied in occasion-setting preparations. In positive occasion-setting, animals learn that a target cue (X) results in a reward outcome (+) only when that cue is accompanied by a contextual feature (A); the same cue presented in absence of this contextual feature remains without consequence (A:X+ / X-). Conversely, in negative occasion-setting, a target cue results in reward, except when accompanied by a specific contextual feature (A:X-/ X+). In both preparations, the contextual feature is thought to hierarchically modulate (positively or negatively) the association between the target cue and the outcome [3, 22]. Strong evidence for this hierarchical modulation is the fact that a contextual feature can simultaneously function as a positive modulator for one target cue and as a negative modulator for another target cue (A:X+ / X-/ A:Y-/ Y+). In such bidirectional occasion-setting procedures all cues and contexts have equal probabilities of reward, therefore, accurate reward predictions cannot be achieved via simple associative summation but instead require hierarchical contextual modulation.

Compared to simple forms of associative learning, the neural bases of context-gated associative predictions (occasion-setting) remain severely understudied. Although recent studies have implicated the orbitofrontal cortex (OFC) in occasion-setting [23–25], most of the behavioral and neurobiological occasion-setting studies have exclusively used male subjects and/or did not explicitly address sex as a biological variable.

Therefore, the purpose of this study was two-fold: 1) investigate sex differences in the (bidirectional) contextual gating of reward prediction and its sensitivity to stress, and 2) investigate sex differences in OFC engagement during the contextual gating of reward predictions. We found that female rats were slower than males to acquire discriminated (context-gated) reward-seeking. However, once established, context-gated performance was more stable in females; it showed less variability within a session and was more resistant to acute stress. Moreover, we showed that this superior contextual control achieved by females was accompanied by an increased activation of the orbitofrontal cortex.

Results

Male and female rats were randomly assigned to one of four Pavlovian discrimination tasks outlined in Figure 1A. In these tasks [adapted from 26], all rats were exposed to two auditory cues (X and Y, 10s) presented within two distinct, alternating, visual contexts (A and noA, 2min). All rats received similar amount of reward, and produced a similar behavioral response (magazine approach). However, the different reward contingencies featured in these tasks presumably promoted the emergence of different associative architectures for reward prediction (Fig. 1B).

Sex differences in the acquisition of context-gated but not context-independent reward predictions.

A Apparatus and general task design. During each session, the local context alternates every 2min between a Ctx.A (houselight on) and Ctx.noA (houselight off). Embedded in these context cycles, brief auditory stimuli (X or Y; 10s) are presented and potentially followed by reward (chocolate milk or banana pellets). B Reward contingencies and corresponding associative structures for each training condition. Ctx-dep. O1: the local context informs the predictive status of the ambiguous reward cues X and Y. Simple discrimination: reward cues are not ambiguous (only X is rewarded) and the local context is irrelevant. No Discrimination: reward cues X and Y are potentially rewarded and the local context is irrelevant. Ctx-dep. O1/O2: Context-dependent discrimination with differential outcomes; in this case, the local context is relevant but may contribute to context-sensitive performance via simple (non-hierarchical) associations (summation of cue- and context-evoked predictions). C Time in port (% cue time) during rewarded and non-rewarded trials across acquisition for male and female rats trained in each task. D Discrimination ratio across acquisition for male and female rats trained in each task. Error bars and error bands indicate ± s.e.m. *: P < 0.05, t-tests

In the Ctx-dep. O1 task, cue X was rewarded only in context A and cue Y was rewarded only in context noA; all rewarded trials resulted in the same outcome (O1). Accurate reward prediction in this task critically relies on the context-informed hierarchical modulation associative predictions (i.e occasion-setting). The Ctx-dep. O1/O2 task featured similar contingencies with the exception that the two rewarded trials (A:X and Y) resulted in two different outcomes (O1 and O2, respectively). This allowed for distinct context-outcome associations (Ctx.A→O1 and Ctx.noA→O2), as well as distinct cue-outcome associations (X→O1 and Y→O2). As a result, context-sensitive performance could potentially be achieved via simple (nonhierarchical) summation process. In the Simple discrimination task, cue X was always rewarded and cue Y was never rewarded, regardless of context. Finally, in the No discrimination task, both cues X and Y were probabilistically rewarded in both contexts. Importantly, the only task that unambiguously engages context-gated reward predictions (or contextual occasion setting) is the first task: context-dependent discrimination with a single outcome (ctx-dep. O1).

Male rats acquire context-regulation over reward-seeking faster than females

Fig.1 C-D displays the course of acquisition of discriminated Pavlovian approach in all four tasks. In the context-dependent discrimination with a single outcome (Ctx dep. O1), Linear Mixed Models (LMM) analysis revealed a significant Trial x Session interaction, (F(17, 181.488) = 73.343, P < 0.001), as rats gradually learned to respond more during rewarded than non-rewarded trials. Critically, the analysis also revealed a significant Sex x Trial x Session interaction (F(17, 181.488) = 4.683, P < 0.001). Post-hoc comparisons revealed that males and females displayed similar levels of responding to rewarded trials (Ps > 0.260), but females maintained higher responding during non-rewarded trials, causing both sexes to differ in this measure from session 40 onwards (Ps < 0.047).

To facilitate comparisons between groups, we calculated for each animal a discrimination ratio, defined as the time in port during rewarded trials divided by the total time in port during all trials (Fig 1D). With the exception of animals trained in the No Discrimination task (for which discrimination between rewarded and non-rewarded trials was impossible by design), all training groups demonstrated a gradual increase in discrimination ratio (main Session effect: No Discr. P = 0.477; all other groups Ps < 0.001). Importantly, only in Ctx-dep.O1 did we detect a main effect of Sex (F(1, 69.876) = 15.185, P < 0.001) and a Sex x Session interaction (F(17, 365.569) = 3.054, P < 0.001). Indeed, acquisition of discriminated performance was faster in males, causing significant sex differences from session 28 onwards (Ps < 0.014). No main effect or interaction with Sex was detected for any other training group (Ps > 0.305).

We next calculated the proportion of male and female rats that successfully acquired discriminated responding during the training period (Fig. 2). The criterion for successful discrimination was defined as a time in port during rewarded trials that exceeded the time in port during non-rewarded trials by at least 50% (Time rew ≥ 1.5* Time nonrew, which corresponds to a discrimination ratio ≥ 0.6) for a minimum of 4 out of 5 consecutive sessions. Rats in the Ctx-dep. discrimination groups had to reach this criterion for both cues to be considered “discriminators” (A:X ≥ 1.5* X and Y ≥ 1.5 A:Y). A significant sex difference in the proportion of discriminators was observed in the Ctx-dep. O1 group (χ2 = 5.35; P = 0.021), as a greater proportion of males reached criterion (Fig. 2 B-C). No sex differences were observed in the proportion of discriminators in any other tasks.

Individual differences in the acquisition of discriminated Pavlovian approach.

A Time in port (% cue time) during the four different trial types (A:X / X / A:Y / Y) at the end of the training period (last 3 days average) for male and female rats trained in each discrimination task. B Discrimination ratio at the end of the training period (last 3 days average) for male and female rats trained in each discrimination task. Dark symbol: discriminator; Grey symbol: non-discriminator. Horizontal bar = group median; Dashed horizontal line represents the indifference level (no discrimination). C Proportion of male and female rats classified as discriminator or non-discriminator in each task. D Speed of acquisition (sessions to criterion) for discriminator rats, in each task. Horizontal bar = group median *: P < 0.05 Males vs. Females, t-tests; §: P < 0.05 chi-square test.

We then analyzed the number of sessions required for rats to reach the discrimination criterion (Fig. 2D). Rats that failed to reach the discrimination criterion were excluded from this analysis. A Kruskal-Wallis one-way ANOVA revealed a significant effect of the behavioral task (H (2) = 87.361, P < 0.001) with different sessions-to-criterion for all the discrimination tasks (Ctx-dep. O1 > Ctx-dep. O1/O2 > Simple; Ps < 0.001, Dunn’s tests). Sex differences in the speed of acquisition (i.e. sessions-to-criterion) were analyzed separately within each task. In Ctx-dep. O1, males were significantly faster to reach the discrimination criterion (T(38) = 2.502, P = 0.017). No sex differences in the speed of acquisition were observed in the other tasks (Ps ≥ 0.175).

These results indicate that female rats were generally slower to acquire contextual control over cue-evoked reward-seeking. This sex difference in the development of discriminated performance was observed only in Ctx-dep. O1, the only task that unambiguously engages contextual modulation of reward predictions. Note that estrous cycle did not affect performance in female discriminators (Fig. S1).

Trial history influences context-gated reward-seeking in male but not female rats

The delivery of reward, while critical for learning, can also interfere with the expression of context-dependent discrimination [27]. Therefore, we investigated to which extent within-session reward history (i.e. prior rewarded or non-rewarded trials) influenced performance in the Ctx-dep. O1 task, in male and female discriminators (Fig. 3). Performance was analyzed over 10 consecutive sessions (sessions 71-80; 800 trials total). The first trial of a session was excluded as it was not preceded by any trial. For the remaining 790 trials (79 trials x 10 sessions), we determined if the trial was preceded by a rewarded or non-rewarded trial. Responding during non-rewarded and rewarded trials was analyzed separately.

Recent rewards disrupt contextual control of reward seeking in male, but not female, rats.

A Time in port (% cue time) during non-rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR) for male and female rats trained in Ctx-dep. O1 task. Inset: increase in responding caused by prior reward for male and female rats (horizontal bar = median difference score). B Time in port during non-rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. C Response probability on non-rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: increase in the probability of responding caused by prior reward for male and female rats (horizontal bar = median difference score). D Response probability on non-rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. E Time in port (% cue time) during rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: difference in responding caused by prior reward for male and female rats. F Time in port during rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. G Response probability on rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: difference in the probability of responding caused by prior reward for male and female rats. H Response probability on rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. Error bars indicate ± s.e.m. * P < 0.05 Males vs. Females; # P < 0.05; ## < 0.01 Prior R vs. prior NR; t-tests

Fig. 3A shows the time in port during non-rewarded trials, as a function of the outcome of the previous trial (prior reward, or non-reward). A RM-ANOVA revealed a main effect of Prior Rewards (F(1, 38) = 27.144; P < 0.001) as prior rewards increased responding during subsequent non-rewarded trials. The analysis showed no main effect of Sex (F(1, 38) = 1.783; P = 0.190), but a marginal Sex x Prior Reward interaction (F(1, 38) = 4.015; P = 0.052), as males were slightly more influenced by prior rewards. The effect of trial history appears to be carried exclusively by the outcome of the previous trial (t-1) (no effect of consecutive prior rewards or non-rewards; Ps ≥ 0.147, Bonferroni t-tests; Fig. 3B).

We then analyzed another metric of conditioned responding: the response probability (i.e. probability subject enters the reward port during cue presentation). We reasoned that this metric might be more prone to detect brief lapses in discrimination performance (false alarms). Fig. 3C shows the probability of responding to a non-rewarded cue as a function of the outcome of the prior trial (prior reward or non-reward). A RM-ANOVA revealed a main effect of Prior Reward (F(1, 38) = 25.607; P < 0.001), and a significant Sex x Prior Reward interaction (F(1, 38) = 9.247; P = 0.004). In males, prior reward increased the probability of responding on subsequent non-rewarded trials (T(38) = 43.752; P < 0.001) while in females, trial history had no effect (T(38) = 1.631; P = 0.209). As a result, males and females differed in their probability of responding to non-rewarded trials when those trials were preceded by reward (T(38) = 4.508; P = 0.040). Here again we found no influence of the number of consecutive past reward or non-rewards (Ps ≥ 0.710, Fig. 3D).

Responding to rewarded cues remained high throughout the sessions and was not influenced by trial history (Ps ≥ 0.113); moreover no main effect or interaction with Sex was observed on these rewarded trials (Ps ≥ 0.127) (Fig. 3E-H). These results indicate that male rats have difficulty inhibiting responding to non-rewarded (contextually irrelevant) cues following a recent reward. In contrast, female rats maintained high discrimination performance throughout the session and recent rewards had little to no effect on their discrimination accuracy.

Acute stress disrupts context-gated reward-seeking in male but not female rats

We then tested the effect of an acute stress (90-min restrain) on discrimination performance (expressed as discrimination ratio), for all training groups (Fig. 4). With the exception of the No Discrimination group, acute stress reduced discrimination accuracy in all tasks (main Stress effect: Ps ≤ 0.004). However, only in the Ctx-dep. O1 task did we observe a significant Stress x Sex interaction F(1, 24) = 5.665; P = 0.026) as only males showed reduced discrimination performance following acute stress (Stress vs. Baseline; Males: T(14) = 5.165; P < 0.001; Females: T(10) = 1.290; P = 0.209). No main effect or interaction with Sex was observed in any other task (Ps ≥ 0.178). Therefore, although acute stress disrupted discrimination performance in all tasks, females trained in Ctx-dep. O1 appear to be protected from this disrupting effect of stress.

Acute stress disrupts context-gated reward seeking in male but not female rats

A-B Time in port (% cue time) during the four different trial types (A:X / X / A:Y / Y) at baseline and after acute stress for male (A) and female (B) rats trained in each discrimination task. C Discrimination ratio at baseline and after acute stress, for male and female rats trained in each discrimination task. D Stress-induced disruption in discrimination performance (stress – baseline) for male and female rats trained in each discrimination task (horizontal bar = median difference score). Error bars indicate ± s.e.m; * P < 0.05; *** P < 0.001, t-tests

Female rats show higher orbitofrontal engagement during context-gated reward predictions

For a subset of animals, we quantified c-Fos-immunoreactive (cfos+) cells in the OFC (medial, ventral and lateral subregions) following regular behavior testing (no-stress condition). We also quantified cfos+ in two control regions, M1J and S1J, where we did not expect task-specific activation. Compared to baseline (homecage controls), behavioral testing increased cfos+ expression in all regions of interest, regardless of task (Fig. 5; homecage c-Fos+ counts are shown but not included in the following statistical analyses). A RM-ANOVA revealed a main effect of Region F(4, 192) = 85.037; P ≤ 0.001) and a significant Task x Sex x Region interaction (F(12, 192) = 1.9; P = 0.036). Follow-up planned contrast analysis compared males and females within each task and brain region. In the Ctx-dep. O1 task, females showed increased c-fos+ expression compared to males, in all OFC subregions (T(14) ≥ 2.387; P ≤ 0.032). No sex differences were found in any other task (Ps ≥ 0.255). Moreover, no sex differences were found in the control regions M1J and S1J, regardless of task (Ps ≥ 0.139).

Female rats show higher orbitofrontal engagement during context-gated reward seeking.

A Experimental design. Brains were collected 2h after the start of the behavioral session. Brain sections were stained for c-Fos (activity marker) and NeuN (general neuronal marker). B Behavioral performance (discrimination ratio) ratio during the final session, before tissue collection for male and female rats trained in each task. C-G c-Fos+ density in the primary jaw motor cortex (C), primary jaw somatosensory cortex (D), medial orbitofrontal cortex (E), ventral orbitofrontal cortex (F), and lateral orbitofrontal cortex (G) for males and females trained in each discrimination task. Error bars indicate ± s.e.m. Gray bands indicate average c-fos+ density for animals who remained in their home cage prior to tissue collection (homecage controls; n=5). *: P < 0.05 Males vs. Females, t-tests.

Discussion

We investigated sex differences in contextual modulation of cue-evoked reward predictions and its neural correlates. Male and female rats were trained on a context-dependent discrimination task (Ctx-dep. O1) in which a contextual feature informs the validity of two reward cues (i.e. bidirectional occasion-setting). Overall, males acquired discriminated, context-gated behavior faster than females. However, once established, contextual control over behavior was more robust in females; it was more stable within a session (less influence of prior reward) and more resistant to acute stress. This superior contextual control achieved by females was accompanied by elevated c-Fos expression in all OFC subregions (medial, ventral and lateral). Critically, no sex differences (behavioral or neurobiological) were observed among animals trained on simpler discrimination tasks that do not engage contextual gating (Simple Discrimination, No discrimination, Ctx-dep. O1/O2).

For Ctx-dep. O1, context is an occasion-setter, i.e. a stimulus that hierarchically modulates the associative strength between a target cue and its outcome. However, our task differs from classic occasion-setting preparations in several ways. Historically, Pavlovian modulation has been studied in preparations in which the modulating stimuli (the occasion-setter) exerts unidirectional modulation, either strengthening (A:X+ / X-) or weakening (A:X-/ X+) a cue-outcome association. In such preparations, the occasion-setter might also establish a direct association (excitatory or inhibitory) with the outcome [1, 28, 29], thereby entangling the contributions of hierarchical modulation and direct stimuli-evoked predictions. To circumvent this issue, occasion-setting preparations commonly introduce a temporal gap (5-30s) between the termination of a discrete occasion-setting stimulus and its target cue. This approach has generally been successful in limiting direct associations between the occasion-setter and the outcome, instead promoting hierarchical modulation processes [30, 31, but see 32]. This temporal gap however introduces a working-memory requirement (animals must remember if the target cue was preceded by the occasion-setter) which complicates results interpretations. Here, we used a bidirectional occasion-setting preparation with a single reward outcome (Ctx-dep. O1) as an alternative approach to investigate contextual modulation [26]. Here, contextual feature (A) positively modulates one target association (A:X+ / X-) and negatively modulates another association (A:Y-/ Y+). Critically, there are no direct informative associations between contextual feature A and the single reward outcome (rewards are equally frequent in the presence or absence of A), so accurate discrimination necessitates hierarchical modulation [22, 33]. Moreover, the contextual feature precedes and overlaps with the target cue, thereby eliminating working-memory requirements. Thus, sex differences observed in Ctx-dep. O1, strongly indicate sex differences in hierarchical (context-informed) associative modulations. Alternatively, males and females might differ in terms of attentional resources. Indeed, unlike Simple Discrimination or No discrimination tasks, Ctx-dep. O1 requires high multimodal attention (animals must simultaneously attend to the visual context and auditory cues). However, no sex differences were observed in context-dependent discrimination with two outcomes (Ctx-dep. O1/O2), a task that shares similar attentional requirements with Ctx-dep. O1, but does not necessarily engage hierarchical contextual modulation (context potentially informing predictions via direct context-outcome associations). Together, these results indicate that the sex differences observed here are not attributable to simple associative, motivational, working-memory, or attentional processes, but are specific to the neurocomputational operations required for the hierarchical, contextual control of behavior.

Sex differences in context-dependent behaviors have previously been reported in fear conditioning preparations where females often display generalized fear responses that evade precise contextual control [34–37]. Elevated baseline anxiety and fear reactivity in females might contribute to this relative failure to regulate fear responses [38, 39, but see 40]. However, we show here that reduced contextual control in females (at least in the early stages of acquisition) is not limited to aversive behaviors but extends to appetitive, reward-seeking behaviors. This suggests a broader, domain-general, delay in the engagement of contextual, top-down regulation processes in female rodents. This is consistent with higher sign-tracking tendencies in female rats [41–45] —another form of appetitive behavior that reflects reduced top-down cognitive control (opposed to goal-tracking behavior) [46–48]. Note that the reduced top-down contextual control displayed by females in our study was limited to delayed acquisition. Provided sufficient training, females actually established more robust contextual control over cue-evoked reward seeking than males.

The biological basis for the delayed acquisition of contextual control in females is unknown. Although estrous cycle had no effect on the expression of context-dependent behavior in female discriminators, fluctuating sex hormones (and resulting neurobiological changes) could have contributed to the slower acquisition of contextual control in females. Indeed, learning the hierarchical contextual rules for reward prediction is a complex and lengthy process that probably requires distributed plasticity across several brain regions including the striatum, amygdala, hippocampus and neocortex [3, 49]. Estrous cycle affects synaptic physiology and plasticity at all these sites via changes in dopamine reuptake [50, 51], neuronal excitability [52–56], spine density [57, 58], gene expression [59, 60], or neurogenesis [61]. This dynamic neurobiological environment might constitute a challenge for females attempting to solve the Ctx-dep. O1 task, and might contribute to their delayed acquisition of contextual control. Interestingly, this dynamic neurobiological environment might also contribute to the more stable performance ultimately achieved by females. Indeed, in deep learning models, the injection of random “noise” in (artificial) neural networks during training is commonly used to build redundancy and increase network performance and stability [62, 63]. This ‘robustness through perturbations’ approach often comes at the expense of learning speed, similar to what we observed in females. This hypothesis of a role of cyclical sex hormones in the trade-off between learning speed and performance stability remains to be confirmed experimentally.

Regardless of the precise mechanism, our results indicate that, compared to male rats, females ultimately achieved more stable contextual control over cued reward-seeking; their behavior remained context-regulated under stress or after recent rewards. This behavioral sex difference was accompanied by neural sex differences. Compared to males, females displayed increased OFC activation during Ctx-dep. O1. Importantly, no behavioral or neural sex differences were observed for any other task. Together, these results strongly implicate the OFC in the sex-biased expression of contextual modulation of reward-seeking. This is consistent with recent studies demonstrating a critical role of the OFC in hierarchical control over cue-evoked behaviors [23–25]. More generally, the OFC is proposed to encode task structure including context-dependent contingencies [64– 66]. Accordingly, the OFC appears critical for decision-making when optimal choice requires knowledge of the broader task structure (i.e. when stimuli cannot be taken at face value) [67, 68]. Although causality remains to be established, our data suggest that the increased orbitofrontal engagement displayed by females might contribute to their resilient contextual regulation of behavior.

The contextual regulation of behavior is at the heart of cognitive control [69]. Deficits in this process are integral to the pathophysiology of numerous neuropsychiatric disorders [4]. Although extrapolation of animal data to humans requires caution, this study might offer insight into the biological sex differences observed in certain disorders. For instance, the initial delay in establishing contextual control observed here in females might contribute to the rapid progression of SUD or PTSD observed in women following initial drug use or trauma [13, 14, 70, 71]. Conversely, the more stable contextual control ultimately achieved by females after extended training observed here might contribute to the better prognosis for women following behavioral and cognitive treatments for these disorders [71–73], and might be due, in part, to increased OFC engagement [74].

Methods

Subjects

Subjects were male and female Long Evans rats (Charles Rivers), 8-10 weeks at arrival. Rats were housed in same-sex pairs in a temperature-controlled vivarium (22 °C) with a 12-hr light/dark cycle; behavioral experiments were conducted during the light cycle. Rats were given free access to water in their home cage for the duration of the experiments. After a week of acclimation to the colony room, rats were mildly food-restricted to maintain ∼95% of age-matched free-feeding weights. All experimental procedures were conducted in accordance with UCSB Institutional Animal Care and Use Committees and the US National Institute of Health guidelines.

Apparatus

Behavioral training was conducted in 12 identical conditioning chambers enclosed in individual sound- and light-attenuating cubicles (Med Associates, St. Albans, VT). Six chambers were dedicated to male subjects and six chambers were dedicated to female subjects; animals of both sexes were trained simultaneously. A recessed reward delivery port was located at the center of the front panel and was connected to a pellet dispenser and a syringe pump for liquid reward delivery (syringe pump located outside of the sound-attenuating cubicle). Subjects’ presence in the reward port was detected by the interruption of an infrared beam. Two ceiling-facing chamber lights located on the front and back internal walls of the cubicle provided diffuse chamber illumination and were used to manipulate the background visual context. Auditory stimuli were delivered via a white noise speaker and a clicker (76dB each), located in the back and front panel, respectively. A fan mounted on the cubicle provided ventilation and low background noise. A computer equipped with Med-PC software (Med Associates) controlled all experimental programs and data collection.

Rewards

Food rewards consisted of a 45mg banana-flavored food pellet (BioServe Dustless Precision Pellets, F0059 formula) or a 0.17mL bolus of sweetened chocolate milk (2 parts chocolate Nesquik® + 1 part water + 5% sucrose w/v) delivered over 3s. These rewards are highly discriminable but equicaloric and were equally preferred by rats in preliminary studies. The identity of the rewarding outcomes was counterbalanced within behavioral tasks and within sexes.

Discrimination Training

Rats were randomly assigned to one of four Pavlovian discrimination conditions: Context dependent discrimination with single outcome (Ctx-dep. O1, n = 30F + 32M), Context dependent discrimination with dual outcome (Ctx-dep. O1/O2, n = 14F + 14M), Simple discrimination (Simple, n = 12F + 22M), or No Discrimination (NoDiscr., n = 8F + 8M). Animals were individually placed in the conditioning chambers for 80-min training sessions 5-6 days/week. During each session, the visual background –the local context– alternated every two minutes between flashing houselights (0.25s on, 0.25s off; Context A) or darkness (Context noA). Within each context phase, two 10-s auditory cues, X or Y (white noise or clicker, counterbalanced), were presented individually, in two separate trials, and were potentially followed by reward. The presentation of each cue within a context phase was selected semi-randomly, so that the identity of the first cue had no incidence on the identity of the second cue (i.e. the same cue could repeat twice within a context phase, so the outcome of the first trial did not inform the outcome of the second trial). Each session consisted of 40 context cycles and 80 cue presentations (i.e. 80 trials) with an average intertrial of 50s ± 20s (rectangular distribution). The reward contingencies in each task are outlined in Figure 1.

Context dependent discrimination with single outcome (Ctx-dep. O1)

In this task, the reward-predictive validity of each cue was informed by the context, so that cue X was rewarded only in Ctx.A and cue Y was rewarded only in Ctx.noA. All rewarded trials resulted in the same reward outcome (O1). Considered individually, both cues and both contexts have equal probabilities of resulting in the same outcome, therefore accurate performance in this task cannot be understood in terms of summed associative strength between cues and contexts. Instead, accurate performance in this task critically relies on contextual modulation (i.e. occasion-setting). During the first 35 training sessions, rats experienced an equal number of rewarded and unrewarded trials (20 trials for each context/cue combination per session). On the 36th session and for the remainder of the experiment, the relative proportion of rewarded trials was reduced (5 A:X+; 15 X-; 15 A:Y-; 5 Y+) in an effort to promote discrimination.

Context dependent discrimination with dual outcome (Ctx-dep. O1/O2)

This task followed similar contingencies as previously described, with the exception that the two rewarded trial types (A:X and noA:Y) resulted in two different food outcomes (O1 and O2, respectively). This has profound implications for the learning processes engaged in the task. The delivery of different outcomes in different contexts allows for distinct context-outcome associations (Ctx.A→O1 and Ctx.noA→O2), as well as distinct cue-outcome associations (X→O1 and Y→O2). These different associative structures could potentially facilitate discrimination performance via simple summation process.

Simple Discrimination

Cue X was always rewarded and cue Y was never rewarded, regardless of the context. Presumably, this training protocol promotes accurate reward prediction via simple associative learning (context-independent predictions).

No Discrimination

Both cues X and Y were probabilistically rewarded in both context (the background context provides no disambiguating information). Presumably, this training protocol promotes reward prediction via simple associative learning (context-independent predictions) although the reward cues retain some ambiguity.

Assessment of Estrous Cycle

Estrous cycle phases were monitored via vaginal cytology for a subset of females (n = 10) over 15 consecutive days (sessions 66-80). Immediately after the behavioral session, the tip of a saline-moistened cotton swab was inserted into the vagina and rotated to dislodge cells from the vaginal wall. Swabs were immediately rolled onto a glass slide, and the samples were preserved with a spray fixative (M-FixTM, MilliporeSigma) without allowing the cells to dry. Slides were then stained with a modified Papanicolau staining procedure as follows: 50% ethyl alcohol, 3 min; tap water, 10 dips (×2); Gill’s hematoxylin 1, 6 min; tap water, 10 dips (×2); Scott’s water, 4 min; tap water, 10 dips (×2); 95% ethyl alcohol, 10 dips (×2); modified orange-greenish 6 (OG-6), 1 min; 95% ethyl alcohol, 10 dips; 95% ethyl alcohol 8 dips; 95% ethyl alcohol 6 dips; modified eosin azure 36 (EA-36), 20 min; 95% ethyl alcohol, 40 dips; 95% ethyl alcohol, 30 dips; 95% ethyl alcohol, 20 dips; 100% ethyl alcohol, 10 dips (×2); xylene, 10 dips (×2); coverslip immediately. Samples were observed with light microscopy (4x). Four criteria were used to determine the specific estrus cycle: 1. cell density, 2. percentage of nucleated epithelial cells, 3. cornified epithelial cells, and 4. Percentage of leukocytes [75–77]. Each sample was independently rated by two trained observers (S.P. and A.M.), blind to the behavioral data. Inter-rater agreement was high (88.2%). Only behavioral sessions for which both raters agreed on the estrous phase were selected for analysis. As a result, 2-4 behavioral sessions were considered for each subject at each phase of the estrous cycle. Behavioral data on these sessions were averaged to express a single data point for each rat at each stage of the estrous cycle. Females that did not show a regular cycle were excluded from the analysis (n = 2).

Acute Restraint Stress

Before testing the effect of acute stress on discrimination performance, the relative proportion of rewarded trials was adjusted in all groups to match the proportion of rewarded trials in the context-dependent discrimination with a single outcome (Ctx-dep. O1). As a result, rewarded cues became less frequent and represented only 1/4 of the trials. For all groups, these rewarded trials were equally distributed across both contexts. After 5-6 sessions of adjustment to these new conditions, we tested the effect of an acute stress on discrimination performance in all training groups. Acute stress was induced by restraining rats in standard Plexiglas rat restraint tubes for 90 min, in a novel brightly lit room. To account for the size differences between male and females, two different sized restraint tubes were used. After placing the rats in the tube, the length of the restrainer was adjusted to immobilize the rat without causing pain or interfering with animals’ breathing. After 90 min of restraint, the rats were returned to their homecage and left undisturbed for 5 min before being placed in the conditioning chambers for a normal behavioral session.

cFos Immunofluorescence

Following stress test, animal were retrained for 5-6 sessions, under standard conditions (no stress; equal number of rewards for all training groups). After their last behavioral session, animals remained in the conditioning chambers for 40 min (120 min total); they were then deeply anesthetized with a lethal dose of Euthasol and perfused transcardially with ice-cold 1x PBS followed by ice-cold 4% PFA in PB. Brains were extracted, post-fixed (4% PFA overnight) and cryoprotected (30% sucrose + 0.1% sodium azide, 1x PBS) until processed for immunostaining. Brains were frozen and sectioned into 40μm with a sliding cryostat (Leica CM1860). Free floating coronal sections were washed for 10 min in PBS, followed by a 2-h incubation in a blocking solution (5% NGS, 0.2% Triton X-100 in 1x PBS) at room temperature. Sections were then incubated with rabbit anti-cFos (1:1000, Cell Signaling, CAT#: 2250) and mouse anti-NeuN (neuron-specific protein; 1:2000 or 1:3000, Novus, CAT#: NBP1-92693) primary antibodies, in the blocking solution, for 20 h a 4°C. Following primary antibody incubation, sections were washed in PBS 3 times for 10 min and then incubated with anti-rabbit Alexa Fluor® 647-conjugated secondary antibody (far-red) (1:250, Jackson ImmunoResearch, CAT#: 111-605-144) and anti-mouse Alexa Fluor® 488-conjugated secondary antibody (green) (1:500, Jackson ImmunoResearch, CAT#: 115-545-003) for 90 min at room temperature. Sections were then washed in PBS three times for 10 min, mounted on slides and immediately coverslipped with a hardset mounting medium (Vector laboratories, CAT#: H-1700). Images were acquired at a 10x magnification, using a Keyence BZ-X fluorescence microscope.

Brain regions of interest for c-Fos positive neurons quantification included the medial, ventral, and lateral orbitofrontal cortex (mOFC, vOFC, and lOFC). Additionally, two control regions were included, for which we did not expect task-specific activation: the jaw primary motor (M1J) and somatosensory cortex (S1J). These regions of interests were manually isolated from the original images, using only the green / NeuN channel (the far-red / c-fos channel was temporarily disabled for blinding purposes). A custom ImageJ script (National Institute of Health) was then used to count c-Fos positive neurons in the selected regions and normalize this count to the surface area of the selected region (c-Fos positive neurons / mm2). For each brain region, c-Fos positive neurons were counted bilaterally from 2-4 sections per animal and a mean count was computed for each animal.

Statistical analysis

Data were collected over three replication cohorts, each composed of approximately equal number of male and female rats. Cue-evoked reward seeking was quantified primarily as the percentage of time a rat spent in the food cup during the last 5 seconds of cue presentation. Discrimination accuracy was expressed as a discrimination ratio, defined as the time in port during rewarded trials divided by the total time in port during all trials. A ratio of 0.5 indicates equal responding during rewarded and non-rewarded trials (i.e. no discrimination) whereas a ratio of 1 indicates responding exclusively during rewarded trials (i.e. perfect discrimination). Time in port and discrimination ratio during acquisition were aggregated in 4-session bins. These longitudinal repeated measures were analyzed with Linear Mixed-Models (LMM) with a Restricted Maximum Likelihood (REML) estimation method. All covariance structures were explored and the best fitting model was determined by the lower Akaike information criterion score [78, 79]. The selection of a covariance structure influenced the degrees of freedom of the model, possibly resulting in non-integer values. The remaining of the analyses consisted generally of mixed-models repeated measures (RM) ANOVAs with Sex and/or Task as a between-subject factor, and Trial history, Stress, or Brain region as within-subject factors. The Geisser-Greenhouse correction was applied when data lacked sphericity, possibly resulting in non-integer degrees of freedom. Post-hoc or planned contrast pairwise comparisons were carried with Bonferroni-corrected t-tests. When appropriate, nonparametric tests were conducted and consisted of the Kruskal-Wallis H test followed by post-hoc Bonferroni-corrected Dunn’s pairwise comparisons. All tests were two-tailed, significance was assessed against a type I error rate of 0.05. Statistical analyses were conducted with SPSS Statistics package (version 28.0.0.0; IBM SPSS). Graphs were generated using GraphPad Prism 9.

Supporting information

eLife Assessment

This valuable manuscript reveals sex differences in bi-conditioning Pavlovian learning and conditional behavior. Males learn hierarchical context-cue-outcome associations more quickly, but females show more stable and robust task performance. These sex differences are related to cellular activation in the orbitofrontal cortex. Although the evidence for the claims is solid, the claim of sex differences in context-dependent discrimination behaviour is not fully supported by the data. Nevertheless, the results will be of interest to many behavioural neuroscientists, particularly those who investigate sex-specific behaviours.

Reviewer #1 (Public Review):

Summary:

Peterson et al., present a series of experiments in which the Pavlovian performance (i.e. time spent at a food cup/port) of male and female rats is assessed in various tasks in which context/cue/outcome relationships are altered. The authors find no sex differences in context-irrelevant tasks and no such differences in tasks in which the context signals that different cues will earn different outcomes. They do find sex differences, however, when a single outcome is given and context cues must be used to ascertain which cue will be rewarded with that outcome (Ctx-dep O1 task). Specifically, they found that males acquired the task faster, but that once acquired, the performance of the task was more resilient in female rats against exposure to a stressor. Finally, they show that these sex differences are reflected in differential rates of c-fos expression in all three subregions of rat OFC, medial, lateral, and ventral, in the sense that it is higher in females than males, and only in the animals subject to the Ctx-dep O1 task in which sex differences were observed.

Strengths:

• Well-written.

• Experiments elegantly designed.

• Robust statistics.

• Behaviour is the main feature of this manuscript, rather than any flashy techniques or fashionable lab methodologies, and luckily the behaviour is done really well.

• For the most part I think the conclusions were well supported, although I do have some slightly different interpretations to the authors in places.

Weaknesses:

1. With regards to the claim (page 4 of pdf), I think I can see what the authors are getting at when they claim "Only Ctx-dep.01 engages context-gated reward predictions", because the same reward is available in each context, and the animal must use contextual information to determine which cue will be rewarded. In other words, it has a discriminative purpose. In Ctx-dep.O1/O2, however, although the context doesn't serve a discriminative purpose in the sense that one cue will always earn a unique outcome, regardless of context, the fact that these cues are differentially rewarded in the different context means that animals may well form context-gated cue-outcome associations (e.g. CtxA-(CS1-O1), CtxnoA-(CS2-O2)). Moreover, the context is informative in this group in telling the animal which cue will be rewarded, even prior to outcome delivery, such that I don't think contextual information will fade to the background of the association and attention be lost to it in the way, say Mackintosh (1975) might predict. Therefore, I don't think this statement is correct.

2. I think the results shown in Figure 1 are very interesting, and well supported by the statistics. It's so nice to see a significant interaction, as so many papers try to report these types of effects without it. However, I do wonder how specific the results are to contextual modulation. That is, should a discriminative discrete cue be used instead of each context (e.g. CS1 indicates CS2 earns O1, CS3 indicates CS4 earns O1), would female rats still be as slow to learn the discrimination?

3. Pages 8-9 of pdf, where the biological basis or the delayed acquisition of contextual control in females is considered, I find this to be written from a place of assuming that what is observed in the males is the default behaviour. That is, although the estrous cycle and its effects on synaptic plasticity/physiology may well account for the results, is there not a similar argument to be made for androgens in males? Perhaps the androgens also somehow alter synaptic plasticity/physiology, leading to their faster speed, reduced performance stability, and increased susceptibility to stress.

4. In addition, the OFC - which is the brain region found to have differential expression of c-fos in males and females in Figure 5 - is not explicitly discussed with regard to the biological mechanisms of differences, which seems odd.

Reviewer #2 (Public Review):

Summary:

A bidirectional occasion-setting design is used to examine sex differences in the contextual modulation of reward-related behaviour. It is shown that females are slower to acquire contextual control over cue-evoked reward seeking. However, once established, the contextual control over behaviour was more robust in female rats (i.e., less within-session variability and greater resistance to stress) and this was also associated with increased OFC activation.

Strengths:

The authors use sophisticated behavioural paradigms to study the hierarchical contextual modulation of behaviour. The behavioural controls are particularly impressive and do, to some extent, support the specificity of the conclusions. The analyses of the behavioural data are also elegant, thoughtful, and rigorous.

Weaknesses:

My primary concern is that the authors' claim of sex differences in context-dependent discrimination behaviour is not fully supported by their data.

First, the basic behavioural effect does not seem to replicate across experiments. The authors first show sex differences in the % time in food port and the discrimination ratio (Figures 1 and 2) such that males show better context-dependent discrimination than females (group ctx-dep O1). However, this difference is not observed in the baseline condition group in the next experiment, which investigates the effect of acute stress on context-gated reward seeking: "In Figure 4, we observe no difference between males versus females in group "ctx-dep O1".

Second, I am not fully convinced by the authors' assertion that the results are specific to the contextual modulation process. The authors' main conclusions are derived from comparing a group trained with the differential outcome procedure (group cxt-dep O1/O2) and a group with the non-differential outcome procedure (group cxt-dep O1). However, importantly, a different number of training sessions was used for ctx-dep O1/O2 and ctx-dep O1. Is it not possible that sex differences could have emerged with additional training in the cxt-dep O1/O2 group? Moreover, the authors also seem to assume that rats are not using a contextual strategy in the context-dep O1/O2 condition (i.e., rats use instead distinct context-outcome associations) but what is the evidence for this? Also, the authors argue that the impact of stress is specific to the hierarchical contextual modulation of behaviour however inspection of Figure 4A suggests that there may also be an effect of stress on the context-dependent O1/O2 group.

I also had some minor issues with how the authors interpreted some of the findings. First, it is shown that recent rewards disrupt contextual control of reward seeking in male, but not female, rats. That is, in males, prior reward increased the probability of responding on subsequent non-rewarded trials but trial history had no effect in females. How do the authors reconcile this finding with the quicker acquisition and better discrimination that is observed in males? It is not evident to me how males can have difficulty inhibiting responding to non-rewarded cues following recent reward yet still show better discrimination throughout training.

Finally, the authors argue that the contextual control over behaviour was more robust in female rats as females show less within-session variability and greater resistance to stress. What evidence is there that the restraint stress procedure causes a similar stress response in both sexes?

Reviewer #3 (Public Review):

Summary:

This manuscript reports an experiment that compared groups of rats acquisition and performance of a Pavlovian bi-conditional discrimination, in which the presence of one cue, A, signals that the presentation of one CS, X, will be followed by a reinforcer and a second CS, Y, will be nonreinforced. Periods of cue A alternated with periods of cue B, which signaled the opposite relationship, cue X is nonreinforced, and cue Y is reinforced. This is a conditional discrimination problem in which the rats learned to approach the food cup in the presence of each CS conditional on the presence of the third background cue. The comparison groups consisted of the same conditional discrimination with the exception that each CS was paired with a different reinforcer. This makes the problem easier to solve as the background is now priming a differential outcome. A third group received simple discrimination training of X reinforced and Y nonreinforced in cues A and B, and the final group was trained with X and Y reinforced on half the trials (no discrimination). The results were clear that the latter two discrimination learning procedures resulted in rapid learning in comparison to the first. Rats required about 3 times as many 4-session blocks to acquire the bi-conditional discrimination than the other two discrimination groups. Within the biconditional discrimination group, female and male rats spent the same amount of time in the food cup during the rewarded CS, but females spent more time in the food cup during CS- than males. The authors interpret this as a deficit in discrimination performance in females on this task and use a measure that exaggerates the difference in CS+ and CS_ responding (a discrimination ratio) to support their point. When tested after acute restraint stress, the male rats spent less time in the food cup during the reinforced CS in comparison to the female rats, but did not lose discrimination performance entirely. The was also some evidence of more fos-positive cells in the orbitofrontal cortex in females, but this difference was of degree.

Overall, I think the authors were successful in documenting performance on the biconditional discrimination task. Showing that it is more difficult to perform than other discriminations is valuable and consistent with the proposal that accurate performance requires encoding of conditional information (which the authors refer to as "context"). There is evidence that female rats spend more time in the food cup during CS-, but I hesitate to agree that this is an important sex difference. There is no cost to spending more time in the food cup during CS- and they spend much less time there than during CS+. Males and females also did not differ in their CS+ responses, suggesting similar levels of learning. A number of factors could contribute to more food cup time in CS-, such as smaller body size and more locomotor activity. The number of food cup entries during CS+ and CS- was not reported here. Nevertheless, I think the manuscript will make a useful contribution to the field and hopefully lead readers to follow up on these types of tasks.

One area for development would be to test the associative properties of the cues controlling the conditional discrimination, can they be shown to have the properties of Pavlovian occasion-setting stimuli? Such work would strengthen the justification/rationale for using the terms "context" and "occasion setter" to refer to these stimuli in this task in the way the authors do in this paper.

Strengths:

- Nicely designed and conducted experiment.

- Documents performance difference by sex.

Weaknesses:

- Overstatement of sex differences.

- Inconsistent, confusing, and possibly misleading use of terms to describe/imply the underlying processes contributing to performance.

Author response:

The following is the authors’ response to the original reviews.

We thank the reviewers for their constructive comments on our manuscript and their appreciation of the results. We provide point-by-point responses bellow. For your convenience we highlight here the main changes to the manuscript.

· More descriptive terminology for the contextual cues (Ctx.A / Ctx.noA is now referred to as LIGHT / DARK).

· Schematic of experiment timeline highlighting the exclusion of non-discriminators following the initial acquisition period. This explains the absence of baseline sex differences post acquisition and clears up some misconceptions about lack of replicability.

· New data (time in port preCS) showing that a prior reward does not cause continued presence in port.

· Several text edits to address all the points raised by the reviewers.

We hope that the editors and reviewers will be satisfied with this revised version and find the strength of the evidence more convincing.

Reviewer #1 (Recommendations For The Authors):

In relation to weaknesses points 1-4 in the public review:

(1) With regards to the claim (page 4 of pdf), I think I can see what the authors are getting at when they claim "Only Ctx-dep.01 engages context-gated reward predictions", because the same reward is available in each context, and the animal must use contextual information to determine which cue will be rewarded. In other words, it has a discriminative purpose. In Ctx-dep.O1/O2, however, although the context doesn't serve a discriminative purpose in the sense that one cue will always earn a unique outcome, regardless of context, the fact that these cues are differentially rewarded in the different context means that animals may well form context-gated cue-outcome associations (e.g. CtxA-(CS1-O1), CtxnoA-(CS2-O2)). Moreover, the context is informative in this group in telling the animal which cue will be rewarded, even prior to outcome delivery, such that I don't think contextual information will fade to the background of the association and attention be lost to it in the way, say Mackintosh (1975) might predict. Therefore, I don't think this statement is correct.

I suggest that the authors refine the statement to be more accurate.

We agree with the reviewer —the context is absolutely relevant for rats trained in the Ctx-dep. O1/O2 task. We have edited the text in several places to make this clear. The question is how (by what mechanism) does the context participate in the control of behavior in this group. The reviewer correctly points out that, just like rats trained in the Ctx-dep. O1 task, rats trained in the Ctx-dep. O1/O2 might have formed context-gated cue-outcome associations. We now clearly acknowledge that in the text.

However, because in this group the two outcomes are always encountered in different contexts, we argue that these rats could also have formed a direct association between the two contexts and the two outcomes. In other words, each context might directly evoke the expectation of a distinct reward outcome (prepare to drink, or prepare to eat). On a given trial, if the cue and context both tend to activate the same outcome representation, the converging cue+context excitation can add up. This would produce a context-sensitive response, but not via hierarchical modulation process (unlike Ctx-dep O1). Arguably, this last associative mechanism is much simpler and might explain why almost all rats in Ctx-dep. O1/O2 group learned the discrimination and at a much faster rate.

Therefore, while rats trained in Ctx-dep O1/O2 might engage a combination of associative processes to achieve context-sensitive behavior (including hierarchical associations), only rats in the Ctx-dep O1 critically and unambiguously rely on hierarchical associations to achieve context-sensitive behavior.

(2) I think the results shown in Figure 1 are very interesting, and well supported by the statistics. It's so nice to see a significant interaction, as so many papers try to report these types of effects without it. However, I do wonder how specific the results are to contextual modulation. That is, should a discriminative discrete cue be used instead of each context (e.g. CS1 indicates CS2 earns O1, CS3 indicates CS4 earns O1), would female rats still be as slow to learn the discrimination?

I am just curious as to whether the authors have thoughts on this.

We have not tested this and are not aware of a paper that examined this question specifically.

However, we would like to point out that in the suggested design (CS1→[CS2→O1]; CS3→[CS4→O1]) the discriminative cues (CS1 and CS3) would almost certainly also acquire substantial reward-predictive value, either because of their direct association with the reward, or via second-order conditioning. This would complicate the interpretation of the results in terms of hierarchical associations. Incorporating non-rewarded presentation of CS1 and CS3 alone (i.e. extinguishing those cues, as is sometimes done in occasion setting experiments) would be one way to reduce the reward expectation evoked by those cues, but this approach has some limitations. Indeed, as mentioned by Rescorla (2006) “During extinction, the net associative strength of a stimulus declines to the level of [a response] threshold, but further decrement stops at that point”. So while extinguished CS1 and CS3 might no longer evoke overt behavioral responses, these cues could retain nonnegligible subthreshold excitatory connection with the US. Individually, these cues might fail to evoke responding but could nonetheless increase responding during the CS1→CS2 trials (or CS3→CS4 trials), via simple summation. (Rescorla, 2006: “the compound of two [extinguished] stimuli has a strength that exceeds the threshold and so evokes responding”).

This type of consideration is precisely why we opted for the behavioral task used in the study. In Ctx-dep. O1, the discriminative stimuli exert opposite effects on the two target cues, which rules out summation effects as a mechanism for context-sensitive behavior.

(3) Pages 8-9 of pdf, where the biological basis or the delayed acquisition of contextual control in females is considered, I find this to be written from a place of assuming that what is observed in the males is the default behaviour. That is, although the estrous cycle and its effects on synaptic plasticity/physiology may well account for the results, is there not a similar argument to be made for androgens in males? Perhaps the androgens also somehow alter synaptic plasticity/physiology, leading to their faster speed, reduced performance stability, and increased susceptibility to stress.

I would like the argument that female behaviour might be the default, and male behaviour the deviation to be considered in the discussion in addition to those already stated.

We regret if we gave the impression that male behavior was the default. The paper is intended to report sex differences but we don’t view either sex as the default. To correct this impression, we have added a few sentences in the discussion to highlight male-hormonal factors as well as non-gonadal genetic factors that might have contributed to the observed sex differences.

(4) In addition, the OFC - which is the brain region found to have differential expression of c-fos in males and females in Figure 5 - is not explicitly discussed with regard to the biological mechanisms of differences, which seems odd.

I suggest OFC be discussed with regard to biological mechanisms of differences.

We added a few sentences in the discussion to i) highlight the parallel between our study and human fMRI studies showing superior OFC activation in females during the regulation of emotional responses, ii) Suggest a potential relationship between the reported sex differences (speed of acquisition, robustness of performance, and OFC activation in context-gated reward prediction), iii) acknowledge our ignorance of the root causes of these sex differences.

We wish we could offer a better answer. We have attempted to offer possible proximal explanations for the observed sex differences, but ultimately our work did not address the root causes of these behavioral and neural sex differences. Therefore we feel that further attempts to explain these differences would be too speculative.

(5) I did wonder if the authors were aware that in the Rescorla-Wagner model, contextual stimuli are thought to summate with discrete cues to enter into the association with the outcome (i.e., the error term is between lambda and sigmaV, with sigmaV the 'summation' of all stimuli present on a trial, including contextual stimuli). Typically, this is not considered much, because the cue itself is so salient and more consistently paired with reward (whereas the ever-present context is often paired with no reward), but nevertheless, it is a part of the association. I'm not sure it's wrong to say that the background circumstances under which events occur are thought to play little role (as in the second sentence of the introduction), but I was wondering if the authors were aware of this fact when they wrote that.

This sentence in the introduction was meant to introduce the distinction between eliciting stimuli and modulating contexts. Admittedly, this paints a naive picture, which we now acknowledge (we hope that the rest of the paper provides more nuance). As pointed out by this reviewer, the context is also a stimulus, and, just like any other stimulus, it is eligible for direct association with an outcome. The possibility for direct context→outcome association is precisely the rational for the Ctx-dep O1/O2 group.

(6) Context-noA - Seems a little confusing for a name, why not just call it context B? NoA appears to imply that nothing happens in A or no outcome is available, whereas this is not always the case.

We debated which terminology to use. We felt that “Context A vs. Context B” should perhaps be reserved to situations where the global context changes (e.g. two different conditioning boxes with different odors, floor texture etc., with proper counterbalancing procedures). We felt that “Context A vs noA” might be more appropriate here, as we are manipulating the local context by introducing (or removing) one single stimulus (the houselight). In this revised version we followed this reviewer’s advice and adopted a more descriptive, and hopefully less confusing, terminology: "Light vs Dark”.

(7) Why is it that in the text the Ctx-dep O1/O2 is explained before simple and no discrimination, but in the Figure Ctx-dep O1/O2 is shown last? These should be consistent.

Thanks for pointing that out. We have switched the order of task description to be consistent with the figures.

(8) Page 6 (of pdf) - could the authors elaborate a little on why or how (or both) the delivery of reward can interfere with the expression of context-dependent discrimination? Do they just mean the performance of discrimination (e.g., animals will sit at the food port longer if there is food there because they are sitting there and eating it, which does not necessarily reflect the expectation of food based on cue presentations?), in which case it is not the discrimination itself that is being interfered with, just the measure of it. Perhaps the authors could elaborate by just inserting a sentence.

We have added a few sentences to discuss this effect.

The first clarification that we can make is that the reduced discrimination performance following reward is not simply due to animals’ continued presence in the reward port. We have added the time pre-cue to Fig. 3 B-F. This measure is not affected by previous reward history, showing that rats are leaving the port between trials.

So what is driving this effect? At this stage, we are agnostic about the mechanism(s) for this effect. Kuchibhotla et al. (2019) —who first reported a similar effect— proposed a model in which recent rewards modify the threshold for behavioral responses (i.e. performance). In this model, a cue might evoke a weak reward prediction but evoke a strong behavioral response if presented after a reward. Additionally, we believe that learning factors might also contribute to the effect reported here. Indeed, the behavioral response on a given trial likely reflects the balance of hierarchical (context-dependent) associations vs. direct associations (Bradfield and Balleine, 2013). Naturally, this balance is dynamic and influenced by trial history. For instance, a Light:X+ trial might increase the value of cue X and promote responding during the following Dark:X- trial. The same logic could be applied to the influence of the context (e.g., Light:X+ trial might promote responding to a subsequent Light:Y- trial). We are currently working on a computational model that captures the dynamic interplay between hierarchical associations and direct associations. We hope that this model will provide some insight into the learning/performance mechanism for the effects reported here. However this computational work is still in the early stages and beyond the scope of the present study.

(9) The lack of effect in the Ctx-dep O1/O2 groups in Figure 4 could be due to a lack of power - the group sizes are a lot smaller for this group than for Ctx-dep O1 where an interaction was detected. I think this should be at least addressed in the discussion (i.e., that this lack of effect is possibly due to less power here, as the effects are in the same direction).

Good point. We now acknowledge this limitation in the text.

Reviewer #2 (Recommendations For The Authors):

(1) Please comment on the failure to replicate the sex differences across experiments. Perhaps this is due to some change in the training procedure that is briefly mentioned in the methods (a reduction in the number of rewarded trials) but it is unclear.

The reviewer correctly observed that Fig. 3-5 do not show sex differences in baseline condition. This is not because of a replication failure, but because non-discriminating subjects were excluded from the experiment at the end of the acquisition period (after 72 training sessions). We now clarify this in the Method and Results section. We also added a schematic of the experiment timeline that highlights the exclusion of non-discriminators at the end of the acquisition period (Fig 1).

On the topic of replicability, the data for Ctx-dep O1 was collected over 3 cohorts (over the course of 2 years) and the sex difference pattern was consistent. For instance, the proportion of discriminators vs. non-discriminators for males and females trained in Ctx-dep O1, showed similar patterns across cohorts (see below).

(2) The design of this experiment makes it possible to analyse whether there is a differential outcome effect (DOE). The DOE would indeed predict better discrimination in group cxt-dep O1/O2 versus cxt-dep O1, which seems to be exactly what the authors observe although between-group statistics are not reported. Inspection of Figure 1 suggests that there may be a DOE in females but not in males. I wonder if the authors might consider reanalysing the data to check this.

Indeed, there is clearly a differential outcome effect. We now point out this DOE in relation to the latency to achieve discrimination criterion (Fig. 2 C-D). Rats in the Ctx-dep. O1/O2 group acquired discrimination (reached criterion) much faster than rats in in the Ctx-dep. O1 group.

Following the reviewer’s suggestion, we provide here the results of targeted ANOVAs (focusing exclusively on Ctx-dep. O1 and Ctx-dep. O1/O2) to investigate a potential sex-dependent effect of DOE (i.e. Sex x Task interactions), see figure below. A three-way ANOVA (Sex x Task x Session) conducted on the discrimination index reveal a main effect of Task (F1, 86 = 173.560, P < 0.001), Session (F2.678, 230.329 = 140.479, P<0.001) and a marginal effect of Sex (F1,86 = 3.929, P = 0.051), but critically no Task x Sex or Task x Sex x Session interaction (P ≥ 0.504). A two-way ANOVA (Sex x Task) conducted on the sessions to criterion revealed a main effect of both factors (Sex F1, 63 = 9.52, P = 0.003; Task F1, 62 = 184.143, P < 0.001) but critically, no Sex x Task interaction (P = 0.233). These results indicate that the use of two different outcomes clearly facilitated the acquisition of context-dependent discrimination (DOE effect), but this effect benefited both sexes equally. We thank the reviewer for recommending this analysis.

Differential outcome effect (DOE) affects males and females equally.

A. Discrimination ratio over the acquisition period. B. trials to criterion. Compared to animals trained with a single outcome (Ctx-dep. O1), the introducing dissociable outcomes for the two type of rewarded trials (Ctx-dep. O1/O2) profoundly facilitated the acquisition of discriminated behavior. This effect benefited both sexes equally.

(3) Some minor points for clarification that the authors may also wish to address:

- Figure 3: is data presented from sessions 71-80 only or for all sessions? I didn't fully follow the explanation offered in the results section.

That’s right. The data presented in Fig. 3 considers only sessions 71-80, in discriminator rats —when performance is globally stable. We have edited the text to make this clearer. These 10 sessions represent a total of 800 trials (=10 session * 80 trials). The first trial of a session what not included in the analysis since it was not preceded by any trial. For the remaining 790 trials (10 session x 79 trials), we examined how the outcome of the past trial (reward or nonrewarded) influenced responding on the next trial. This large sample size (790 trials / rat) was required to ensure that enough data was collected for each possible trial history scenario.

- The authors argue that females are protected from the disrupting effect of stress. It might be useful if the authors offer further explanation as to what they mean by "protected".

By “protected”, we simply mean “less sensitive”. We have reworded this sentence in that way. We do not claim to have an understanding of the precise mechanism for this sex dependent effect (although our data point to a possible role of the OFC).

- The authors state that "delivery of reward, while critical for learning, can also interfere with the expression of context-dependent discrimination". This statement should be explained in further detail. For instance, why should reward delivery specifically impair context-dependent discrimination but not other forms of discrimination?

We have reworded this sentence to be more inclusive. Indeed, delivery of reward also interferes with other forms of discrimination, particularly when discrimination performance is not yet optimal. We have also added a paragraph to discuss the possible mechanisms by which reward might interfere with discrimination performance in our task.

Reviewer #3 (Recommendations For The Authors):

I do not suggest additional experiments, but I do hope you continue the behavioral work to characterize what is being learned in the task. I think the approach is promising. I would suggest reporting the % time in port and port entries for the entire CS. There is no justification for only analyzing the response in the last 5s.

We thank the reviewer for the encouragement.

We opted to focus on the time in port for two main reasons:

(1) This measure is relatively consistent across the two different reward outcomes (unlike the rate of port entries). Indeed, consistent with prior studies (Delamater et al., 2017), we observed that the type of reward (solid or liquid) influences the topography of the anticipatory magazine-directed behavior. Specifically, cues paired with pellets elicited significantly more port entries than cues paired with chocolate milk. The opposite pattern was observed for time in port --cues paired with chocolate milk elicited more sustained time in port compared to cues paired with pellets (see figure below). While these measures (port entries and time in port) show opposite bias for the two possible outcomes, the size of this bias is much smaller for the time in port (Cohen’s d effect size: port entries: 1.41; time in port: 0.62). As a result, the discrimination ratio calculated from Time in port is consistent across the two outcomes (P = 0.078; effect size: 0.07), which is not the case for the discrimination ratio calculated from port entries (P = 0.007; effect size 0.32 see figure below).

(2) Unlike the rate of port entries, the time in port shows monotonic increase during training in these tasks. Indeed, we observed here and in past work (Keiflin et al., 2019), that the rate of port entries initially increases with training, but then slightly decreases; particularly for cues paired with liquid reward. In contrast, the time in port continues to increase, or remains high, with extended training. This is easy to understand if we consider the extreme case of a hypothetical rat that might enter the port once upon cue presentation and maintain continued presence in port for the whole cue duration. This rat would have a relatively low rate of port entry (a single port entry per trial) but a high time in port.

This is not to say that the rate of port entries is not a valid measure overall (we have used, and continue to use, this metric in other preparations). However, for the reasons explained above, we believe that the time in port is a better metric for reward anticipation in this specific study.

Moreover, we chose to focus our analysis on the last 5s of the cue because that’s when anticipatory food cup behavior is more reliably observed (in our preparation >2/3 of the total time in port in occurs during the last 5s of the cue) and less contaminated by orienting behaviors (Holland, 1977, 1980, 2000). For these reasons, analysis of the last portion of the cue is relatively common in Pavlovian anticipatory approach preparations (El-Amamy and Holland, 2007; Olshavsky et al., 2013; Esber et al., 2015; Holland, 2016a, 2016b; Schiffino and Holland, 2016; Gardner et al., 2017; Sharpe et al., 2021; Maes et al., 2020; Sharpe et al., 2020; Siemian et al., 2021; Kang et al., 2021). Reporting time in port during the same cue epoch facilitates comparisons between these studies.

We have edited the text in the Method section to provide a brief justification for focusing our analyses on this cue epoch.

Outcome identity influences the topography of the conditioned response.

A-C: Conditioned responding expressed as the number of port entries per trial (A) or time in port per trials (C) for rats trained in the simple discrimination task with a chocolate milk reward (n= 19) or a sucrose pellet (n = 16). Data show the average of the last three 3 sessions. Compared to chocolate milk, pellets tend to produce more port entries. Conversely, chocolate milk tend to produce more time in port. However the magnitude of this bias is smaller for the Time in port. C-D: discrimination ratio calculate from the number of port entries (C) or the time in port (D); the latter is not affected by the outcome identity. *P<0.05; **P<0.01; ***P<0.001 T tests.

The inconsistent use of terms is distracting throughout the paper. Is it discriminated or context-gated? Please provide a definition of your terms and then use them consistently. Is it a discriminative stimulus, a context, or an occasion setter? These all imply slightly different things and it would help the reader if you just used one term throughout the paper.

Thanks for pointing that out. We have added a definition for “context-gated” and edited the text to keep the terminology consistent when appropriate. The words “discrimination”/”discriminated” still appear in the manuscript but without implying a mechanism (all tasks are variations of Pavlovian discrimination; the rats discriminating between rewarded and non-rewarded trials).

As mentioned by this reviewer, the terms “context” and “occasion setter” are not synonymous. Therefore these terms still appear in the manuscript to refer to different concepts (e.g. in our task the visual stimulus is a context for all rats; this context acts as an occasion setter only for some rats).

Minor:

Intro, 2nd PP: "autism". This is abbreviated in the abstract but spelled out here. I suggest not abbreviating in the abstract and introducing abbreviations here, as you do with PTSD.

Fixed as suggested

Have deficits in contextual modulation been distinguished from potential deficits in binary associative learning in autism, PTSD, and substance use disorders? This is implied, but there are no citations provided.

We provide a list of references showing deficits in contextual modulation in these disorders.

This does not mean that these disorders are reducible to deficits in contextual modulation and it does not exclude other forms of deficits in those disorders --including alterations in certain aspects of binary associative learning.

"In positive occasion-setting, animals learn that a target cue (X) results in a reward outcome (+) only when that cue is accompanied by a contextual feature (A); the same cue presented in absence of this contextual feature remains without consequence (A:X+ / X-)." - there are words missing in this sentence.

We apologize but we fail identify the missing word(s). Perhaps the reviewer could be more specific and we will be happy to edit the sentence as needed.

What is a contextual feature, is this redundant or can you provide a specific definition?

We use the terminology “feature” and “target” as these are the standard terms in the description of occasion setting preparations (one stimulus, “the feature”, sets the occasion for responding –or not responding- to the “target” cue). By contextual feature, we meant that in this specific example the context was the feature. We have clarified this in the text. We believe that these terms are not redundant. Indeed, the context is not always a feature, and a feature is not necessarily a context (phasic cues can serve as “features”).

Can you provide some background on studies of sex differences in simple associative learning? You imply these have been much more thoroughly studied than conditional discriminations.

We added a few references as suggested.

What is the rationale for studying stress?

Stressful life events exacerbate several mental illnesses, potentially by impacting cognitive functions.

Although the (sex-dependent) effects of stress on some cognitive function are well established (e.g. working memory, selective attention, spatial navigation), the effect of stress on contextual modulation (a core dysfunction in certain mental illnesses) --and the possible sex-differences in this effect-- had not been formally tested. We added a few sentences in the results section (at the beginning of the stress section) to remind the reminder of why we tested the effect of stress in this task.

Method/Results:

Cues are not counterbalanced; the feature is visual and targets are auditory - this should be noted as a limitation in the discussion section.

We now acknowledge this limitation in the discussion. Moreover we believe that the new terminology for the context —Light vs Dark— (instead of A vs. noA in the original version) makes it abundantly clear that the “context” is this study was always visual.

Summation is invoked to describe the discrimination with different outcomes, how is summation happening? This is not described. Perhaps incorporate the literature on conditional discriminations with differential outcomes (the "differential outcomes effect").

We have edited the Result + Discussion section to clarify how summation might contribute to discrimination with different outcomes. We have also added references for the DOE in this task.

The stress effect is confounded with test order; comparing stress vs. baseline.

Sorry we don’t understand this point. The “baseline” refers to the animal’s performance on the last training session before the acute stress manipulation (we have edited the text to make this clear). Animals are first trained in the task and then we examine how stress alters their performance in this learned task. We don’t see how this could induce a test order confound.

Throughout the results section, it would be helpful to have the number of animals reported for each analysis.

The number of animals for each part of the experiment is now reported in the text, as well as in the figures.

Discussion:

"For Ctx-dep. O1, context is an occasion-setter, i.e. a stimulus that hierarchically modulates the associative strength between a target cue and its outcome." This is inaccurate. Occasion setters do not change or modulate the associative strength of a target cue. They modulate whether excitation or inhibition is expressed.

We reworded the sentence as suggested: “For Ctx-dep. O1, context is an occasion-setter, i.e. a stimulus that modulates the response to a target cue”.

"Together, these results indicate that the sex differences observed here are not attributable to simple associative, motivational, working-memory, or attentional processes, but are specific to the neurocomputational operations required for the hierarchical, contextual control of behavior." It should be noted here that the difference is one of degree, a quantitative difference, but not a difference in the qualitative features of the process.

"Regardless of the precise mechanism, our results indicate that, compared to male rats, females ultimately achieved more stable contextual control over cued reward-seeking; their behavior remained context-regulated under stress or after recent rewards." Again this is a matter of degree.

We absolutely agree. All the sex-difference reported here are a matter of degree. In the framework of McCarthy et al. (2012) the reported effects are type 2 or type 3 sex differences, not type 1 sexual dimorphism. We made a few edits in the Discussion to clarify this point.

Procedure:

Please clarify the percentage of trials that were reinforced in the No Discrimination group.

From session 1-32 (acquisition period), 50% of the trials were reinforced. Following this acquisition period, only 25% of the trials were reinforced to match all the other groups. We have edited the method section to clarify this point.

Please provide the dimensions of the restraint tubes and the model number if available.

This information is now included.

References

Bradfield LA, Balleine BW (2013) Hierarchical and binary associations compete for behavioral control during instrumental biconditional discrimination. J Exp Psychol Anim Behav Process 39:2–13.

Delamater AR, Garr E, Lawrence S, Whitlow JW (2017) Elemental, configural, and occasion setting mechanisms in biconditional and patterning discriminations. Behav Processes 137:40–52.

El-Amamy H, Holland PC (2007) Dissociable effects of disconnecting amygdala central nucleus from the ventral tegmental area or substantia nigra on learned orienting and incentive motivation. Eur J Neurosci 25:1557–1567.

Esber GR, Torres-Tristani K, Holland PC (2015) Amygdalo-striatal interaction in the enhancement of stimulus salience in associative learning. Behav Neurosci 129:87–95.

Gardner MPH, Conroy JS, Shaham MH, Styer CV, Schoenbaum G (2017) Lateral Orbitofrontal Inactivation Dissociates Devaluation-Sensitive Behavior and Economic Choice. Neuron 96:1192–1203.e4.

Holland PC (1977) Conditioned stimulus as a determinant of the form of the Pavlovian conditioned response. J Exp Psychol Anim Behav Process 3:77–104.

Holland PC (1980) CS-US interval as a determinant of the form of Pavlovian appetitive conditioned responses. J Exp Psychol Anim Behav Process 6:155–174.

Holland PC (2000) Trial and intertrial durations in appetitive conditioning in rats. Anim Learn Behav 28:121–135.

Holland PC (2016a) Enhancing second-order conditioning with lesions of the basolateral amygdala. Behav Neurosci 130:176–181.

Holland PC (2016b) Effects of amygdala lesions on overexpectation phenomena in food cup approach and autoshaping procedures. Behav Neurosci 130:357–375.

Kang M, Reverte I, Volz S, Kaufman K, Fevola S, Matarazzo A, Alhazmi FH, Marquez I, Iordanova MD, Esber GR (2021) Agency rescues competition for credit assignment among predictive cues from adverse learning conditions. Sci Rep 11:16187.

Keiflin R, Pribut HJ, Shah NB, Janak PH (2019) Ventral tegmental dopamine neurons participate in reward identity predictions. Curr Biol 29:93–103.e3.

Kuchibhotla KV, Hindmarsh Sten T, Papadoyannis ES, Elnozahy S, Fogelson KA, Kumar R, Boubenec Y, Holland PC, Ostojic S, Froemke RC (2019) Dissociating task acquisition from expression during learning reveals latent knowledge. Nat Commun 10:2151.

Maes EJP, Sharpe MJ, Usypchuk AA, Lozzi M, Chang CY, Gardner MPH, Schoenbaum G, Iordanova MD (2020) Causal evidence supporting the proposal that dopamine transients function as temporal difference prediction errors. Nat Neurosci 23:176–178.

McCarthy MM, Arnold AP, Ball GF, Blaustein JD, De Vries GJ (2012) Sex differences in the brain: the not so inconvenient truth. J Neurosci 32:2241–2247.

Olshavsky ME, Song BJ, Powell DJ, Jones CE, Monfils M-H, Lee HJ (2013) Updating appetitive memory during reconsolidation window: critical role of cue-directed behavior and amygdala central nucleus. Front Behav Neurosci 7:186.

Rescorla RA (2006) Deepened extinction from compound stimulus presentation. J Exp Psychol Anim Behav Process 32:135–144.

Schiffino FL, Holland PC (2016) Secondary visual cortex is critical to the expression of surprise-induced enhancements in cue associability in rats. Eur J Neurosci 44:1870–1877.

Sharpe MJ, Batchelor HM, Mueller LE, Gardner MPH, Schoenbaum G (2021) Past experience shapes the neural circuits recruited for future learning. Nat Neurosci 24:391–400.

Sharpe MJ, Batchelor HM, Mueller LE, Yun Chang C, Maes EJP, Niv Y, Schoenbaum G (2020) Dopamine transients do not act as model-free prediction errors during associative learning. Nat Commun 11:106.

Siemian JN, Arenivar MA, Sarsfield S, Borja CB, Russell CN, Aponte Y (2021) Lateral hypothalamic LEPR neurons drive appetitive but not consummatory behaviors. Cell Rep 36:109615.

Version 2

Abstract

Animals, including humans, rely on contextual information to interpret ambiguous stimuli. Impaired context processing is a hallmark of several neuropsychiatric disorders, including schizophrenia, autism spectrum disorders, post-traumatic stress disorder, and addiction. While sex differences in the prevalence and manifestations of these disorders are well established, potential sex differences in context processing remain uncertain. Here we examined sex differences in the contextual control over cue-evoked reward seeking and its neural correlates, in rats. Male and female rats were trained in a bidirectional occasion-setting preparation in which the validity of two auditory reward-predictive cues was informed by the presence, or absence, of a visual contextual feature (LIGHT: X+ / DARK: X− / LIGHT: Y− / DARK: Y+). Females were significantly slower to acquire contextual control over cue-evoked reward seeking. However, once established, the contextual control over behavior was more robust in female rats; it showed less within-session variability (less influence of prior reward) and greater resistance to acute stress. This superior contextual control achieved by females was accompanied by an increased activation of the orbitofrontal cortex compared to males. Critically, these behavioral and neural sex differences were specific to the contextual modulation process and not observed in simple, context-independent, reward prediction tasks. These results indicate a sex-biased trade-off between the speed of acquisition and the robustness of performance in the contextual modulation of cued reward seeking. The different distribution of sexes along the fast learning ↔ steady performance continuum might reflect different levels of engagement of the orbitofrontal cortex, and might have implications for our understanding of sex differences in psychiatric disorders.

Introduction

Pavlovian learning is often conceived as the formation of a binary association between two events, where event #1 (the stimulus) evokes the anticipation of event #2 (the outcome). Under this naive definition, the background circumstances (the context) in which those events occur play little role. However, this simple form of learning is often augmented by higher level cognitive systems that allow animals (including humans) to learn and retrieve context specific associations (e.g., the word “shot” might evoke different representations in a bar vs. a vaccine clinic) [1–3]. This process is referred to as “contextual-gating” (the context acting as a “gatekeeper” in the selection of the appropriate associative memory).

Deficits in contextual gating are observed in several neuropsychiatric disorders including schizophrenia, autism spectrum disorder (ASD), post-traumatic stress disorders (PTSD), and substance use disorders (SUD) [4–12], and may contribute to the intrusive thoughts and maladaptive (situation-inappropriate) behaviors that characterize these disorders. Importantly, the prevalence and symptomatology of these disorders vary between sexes. For example, the incidence of schizophrenia and ASD is higher in men [13] while women are more at risk of developing SUD or PTSD following initial drug use or trauma [14, 15]. This disparity might be due, in part, to subtle sex differences in executive strategies [16–19]. Stress, a contributing factor in all aforementioned disorders, can further exacerbate these sex differences, resulting in sex-specific cognitive vulnerabilities [18–23]. Sex differences in executive functions and sensitivity to stress have been reproduced in animal models using behavioral assays of attention, impulsivity, spatial navigation, and working memory [17, 24–26]. Whether these sex differences extend to the contextual gating of associative predictions remains unknown.

In animal models, the contextual gating of associative predictions can be studied in occasion-setting preparations. In positive occasion-setting, animals learn that a “target” cue (X) results in a reward outcome (+) only when that cue is accompanied by a specific “feature” such as a specific context (A); the same cue presented in absence of this contextual feature remains without consequence (A:X+ / X-). Conversely, in negative occasion-setting, a target cue results in reward, except when accompanied by a specific feature (A:X- / X+). In both preparations, the contextual feature is thought to hierarchically modulate (positively or negatively) the association between the target cue and the outcome [3, 27]. Strong evidence for this hierarchical modulation is the fact that a contextual feature can simultaneously function as a positive modulator for one target cue and as a negative modulator for another target cue (A:X+ / X- / A:Y- / Y+). In such bidirectional occasion-setting procedures all cues and contexts have equal probabilities of reward, therefore, accurate reward predictions cannot be achieved via simple associative summation but instead require the context-informed hierarchical gating of associative predictions (i.e. context-gated associative predictions).

Compared to simple forms of associative learning, the neural bases of context-gated associative predictions (or occasion-setting) remain severely understudied. Although recent studies have implicated the orbitofrontal cortex (OFC) in occasion-setting [28–30], most of the behavioral and neurobiological occasion-setting studies have exclusively used male subjects and/or did not explicitly address sex as a biological variable.

Therefore, the purpose of this study was two-fold:1) investigate sex differences in the (bidirectional) contextual gating of reward prediction and its sensitivity to stress, and 2) investigate sex differences in OFC engagement during the contextual gating of reward predictions. We found that female rats were slower than males to acquire context-gated reward-seeking. However, once established, context-gated performance was more stable in females; it showed less variability within a session and was more resistant to acute stress. Moreover, we showed that this superior contextual control achieved by females was accompanied by an increased activation of the orbitofrontal cortex.

Results

Male and female rats were randomly assigned to one of four Pavlovian discrimination tasks outlined in Figure 1. In these tasks [adapted from 31], all rats were exposed to two auditory cues (X and Y, 10s) presented within two distinct, alternating, visual contexts (LIGHT and DARK; 2min). All rats received similar amount of reward, and produced a similar behavioral response (magazine approach). However, the different reward contingencies featured in these tasks presumably promoted the emergence of different associative architectures for reward prediction (Fig. 1B).

Sex differences in the acquisition of context-gated but not context-independent reward predictions.

(A) Apparatus and general task design. During each session, the local context alternates every 2min between a Ctx. LIGHT (houselight on) and Ctx. DARK (houselight off). Embedded in these context cycles, brief auditory stimuli (X or Y; 10s) are presented and potentially followed by reward (chocolate milk or banana pellets). (B) Reward contingencies and corresponding associative structures for each training condition. Ctx-dep. O1: the local context informs (hierarchically “gates”) the predictive status of the ambiguous reward cues X and Y. Simple discrimination: reward cues are not ambiguous (only X is rewarded) and the local context is irrelevant. No Discrimination: reward cues X and Y are potentially rewarded and the local context is irrelevant. Ctx-dep. O1/O2: Context-dependent discrimination with differential outcomes; in this case, the local context is relevant but may contribute to context-sensitive performance via simple (non-hierarchical) associations (summation of cue- and context-evoked predictions). (C) Experimental timeline. (D) Time in port (% cue time) during rewarded and non-rewarded trials across acquisition for male and female rats trained in each task. (E) Discrimination ratio across acquisition for male and female rats trained in each task. Error bars and error bands indicate ± s.e.m. *: P < 0.05, t-tests

In the Ctx-dep. O1 task, cue X was rewarded only in the LIGHT context and cue Y was rewarded only in DARK context; all rewarded trials resulted in the same outcome (O1). Accurate reward prediction in this task critically relies on the context-informed hierarchical modulation associative predictions (i.e occasion-setting). In the Simple discrimination task, cue X was always rewarded and cue Y was never rewarded, regardless of context. In the No discrimination task both cues X and Y were probabilistically rewarded in both contexts. Finally, the Ctx-dep. O1/O2 task featured similar contingencies as Ctx-dep. O1, with the exception that the two rewarded trials (LIGHT: X and DARK: Y) resulted in two different outcomes (O1 and O2, respectively). In this case, not only might the visual context promote the hierarchical modulation of associative predictions (as in Ctx-dep. O1), but the context might also contribute to context-sensitive performance via simple (non-hierarchical) associations. Indeed, the use of differential outcomes allows for distinct context-outcome associations (LIGHT→O1 and DARK→O2), as well as distinct cue-outcome associations (X→O1 and Y→O2). On a given trial, if the cue and context tend to activate the same outcome representation, the converging cue-evoked and context-evoked excitations can add up. As a result, context-sensitive performance could potentially be achieved via simple (nonhierarchical) summation process. Therefore, while rats trained in Ctx-dep. O1/O2 might engage a combination of associative processes to achieve context-sensitive behavior (including hierarchical and nonhierarchical associations), only rats trained in the Ctx-dep. O1 task critically and unambiguously rely on hierarchical associations to achieve context-sensitive behavior.

Male rats acquire context-regulation over reward-seeking faster than females

Fig.1 C-D displays the course of acquisition of discriminated Pavlovian approach in all four tasks (Ctx-dep. O1 n = 30F + 32M; Simple discrimination n = 12F + 22M; No discrimination n = 8F + 8M; Ctx-dep. O1/O2 n = 14F + 14M). In the context-dependent discrimination with a single outcome (Ctx dep. O1), Linear Mixed Models (LMM) analysis revealed a significant Trial x Session interaction, (F(17, 181.488) = 73.343, P < 0.001), as rats gradually learned to respond more during rewarded than non-rewarded trials. Critically, the analysis also revealed a significant Sex x Trial x Session interaction (F(17, 181.488) = 4.683, P < 0.001). Post-hoc comparisons revealed that males and females displayed similar levels of responding to rewarded trials (Ps > 0.260), but females maintained higher responding during non-rewarded trials, causing both sexes to differ in this measure from session 40 onwards (Ps < 0.047).

To facilitate comparisons between groups, we calculated for each animal a discrimination ratio, defined as the time in port during rewarded trials divided by the total time in port during all trials (Fig 1D). With the exception of animals trained in the No Discrimination task (for which discrimination between rewarded and non-rewarded trials was impossible by design), all training groups demonstrated a gradual increase in discrimination ratio (main Session effect: No Discr. P = 0.477; all other groups Ps < 0.001). Importantly, only in Ctx-dep. O1 did we detect a main effect of Sex (F(1, 69.876) = 15.185, P < 0.001) and a Sex x Session interaction (F(17, 365.569) = 3.054, P < 0.001). Indeed, acquisition of discriminated performance was faster in males, causing significant sex differences from session 28 onwards (Ps < 0.014). No main effect or interaction with Sex was detected for any other training group (Ps > 0.305).

We next calculated the proportion of male and female rats that successfully acquired discriminated responding during the training period (Fig. 2). The criterion for successful discrimination was defined as a time in port during rewarded trials that exceeded the time in port during non-rewarded trials by at least 50% (Time rew ≥ 1.5* Time nonrew, which corresponds to a discrimination ratio ≥ 0.6) for a minimum of 4 out of 5 consecutive sessions. Rats in the Ctx-dep. discrimination groups had to reach this criterion for both cues to be considered “discriminators” (LIGHT:X ≥ 1.5* DARK:X and DARK:Y ≥ 1.5* LIGHT:Y). A significant sex difference in the proportion of discriminators was observed in the Ctx-dep. O1 group (χ2 = 5.35; P = 0.021), as a greater proportion of males reached criterion (Fig. 2 B-C). No sex differences were observed in the proportion of discriminators in any other tasks. We then analyzed the number of sessions required for rats to reach the discrimination criterion (Fig. 2D). Rats that failed to reach the discrimination criterion were excluded from this analysis. A Kruskal-Wallis one-way ANOVA revealed a significant effect of the behavioral task (H (2) = 87.361, P < 0.001) with different sessions-to-criterion for all the discrimination tasks (Ctx-dep. O1 > Ctx-dep. O1/O2 > Simple; Ps < 0.001, Dunn’s tests). Note how the introduction of differential outcomes profoundly facilitated the acquisition of context-dependent discrimination (sessions to criterion: 47.8 ± 1.7 s.e.m. for Ctx-dep. O1 vs. 17.26 ± 1.6 s.e.m. for Ctx-dep. O1/O2). This is consistent with a prior study showing a similar differential outcome effect in this task [31].

Individual differences in the acquisition of discriminated Pavlovian approach.

(A) Time in port (% cue time) during the four different trial types (LIGHT: X / DARK: X / LIGHT: Y / DARK: Y) at the end of the training period (last 3 days average) for male and female rats trained in each discrimination task. (B) Discrimination ratio at the end of the training period (last 3 days average) for male and female rats trained in each discrimination task. Dark symbol: discriminator; Grey symbol: non-discriminator. Horizontal bar = group median; Dashed horizontal line represents the indifference level (no discrimination). (C) Proportion of male and female rats classified as discriminator or non-discriminator in each task. D Speed of acquisition (sessions to criterion) for discriminator rats, in each task. Horizontal bar = group median *: P < 0.05 Males vs. Females, t-tests; §: P < 0.05 chi-square test.

Sex differences in the speed of acquisition (i.e. sessions-to-criterion) were analyzed separately within each task. In Ctx-dep. O1, males were significantly faster to reach the discrimination criterion (T(38) = 2.502, P = 0.017). No sex differences in the speed of acquisition were observed in the other tasks (Ps ≥ 0.175).

These results indicate that female rats were generally slower to acquire contextual control over cue-evoked reward-seeking. This sex difference in the development of discriminated performance was observed only in Ctx-dep. O1, the only task that unambiguously engages contextual modulation of reward predictions. Note that estrous cycle did not affect performance in female discriminators (Fig. S1). Rats that failed to reach the discrimination criterion at the end of the acquisition period were excluded from further experimental procedures (with the exception of rats trained in the No Discrimination group).

Trial history influences context-gated reward-seeking in male but not female rats

The delivery of reward, while critical for learning, can also interfere with the expression of discrimination behavior [32]. Therefore, we investigated to which extent within-session reward history (i.e. prior rewarded or non-rewarded trials) influenced performance in the Ctx-dep. O1 task, in male (n=25) and female (n=15) discriminators (Fig. 3). Performance was analyzed over 10 consecutive sessions (sessions 71-80; 80 trials/ session; 800 trials total), a period during which discrimination performance was globally stable. The first trial of a session was excluded as it was not preceded by any trial. For the remaining 790 trials (10 sessions x 79 trials), we determined if the trial was preceded by a rewarded or non-rewarded trial. This large sample size is required to ensure that enough data was collected for each possible trial history scenario. Responding during non-rewarded and rewarded trials was analyzed separately.

Recent rewards disrupt contextual control of reward seeking in male, but not female, rats.

(A) Time in port (% cue time) during non-rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR) for male and female rats trained in Ctx-dep. O1 task. Inset: increase in responding caused by prior reward for male and female rats (horizontal bar = median difference score). (B) Time in port during non-rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. (C) Response probability on non-rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: increase in the probability of responding caused by prior reward for male and female rats (horizontal bar = median difference score). (D) Response probability on non-rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. (E) Time in port (% cue time) during rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: difference in responding caused by prior reward for male and female rats. (F) Time in port during rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. (G) Response probability on rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: difference in the probability of responding caused by prior reward for male and female rats. (H) Response probability on rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. Error bars indicate ± s.e.m. * P < 0.05 Males vs. Females; # P < 0.05; ## < 0.01 Prior R vs. prior NR; t-tests

Fig. 3A shows the time in port during non-rewarded trials, as a function of the outcome of the previous trial (prior reward, or non-reward). A RM-ANOVA revealed a main effect of Prior Rewards (F(1, 38) = 27.144; P < 0.001) as prior rewards increased responding during subsequent non-rewarded trials. The analysis showed no main effect of Sex (F(1, 38) = 1.783; P = 0.190), but a marginal Sex x Prior Reward interaction (F(1, 38) = 4.015; P = 0.052), as males were slightly more influenced by prior rewards. The effect of trial history appears to be carried exclusively by the outcome of the previous trial (t-1) (no effect of consecutive prior rewards or non-rewards; Ps ≥ 0.147, Bonferroni t-tests; Fig. 3B).

We then analyzed another metric of conditioned responding: the response probability (i.e. probability subject enters the reward port during cue presentation). We reasoned that this metric might be more prone to detect brief lapses in discrimination performance (false alarms). Fig. 3C shows the probability of responding to a non-rewarded cue as a function of the outcome of the prior trial (prior reward or non-reward). A RM-ANOVA revealed a main effect of Prior Reward (F(1, 38) = 25.607; P < 0.001), and a significant Sex x Prior Reward interaction (F(1, 38) = 9.247; P = 0.004). In males, prior reward increased the probability of responding on subsequent non-rewarded trials (T(38) = 43.752; P < 0.001) while in females, trial history had no effect (T(38) = 1.631; P = 0.209). As a result, males and females differed in their probability of responding to non-rewarded trials when those trials were preceded by reward (T(38) = 4.508; P = 0.040). Here again we found no influence of the number of consecutive past reward or non-rewards (Ps ≥ 0.710, Fig. 3D).

Responding to rewarded cues remained high throughout the sessions and was not influenced by trial history (Ps ≥ 0.113); moreover no main effect or interaction with Sex was observed on these rewarded trials (Ps ≥ 0.127) (Fig. 3E-H).

These results indicate that male rats have difficulty inhibiting responding to non-rewarded (contextually irrelevant) cues following a recent reward. In contrast, female rats maintained high discrimination performance throughout the session and recent rewards had little to no effect on their discrimination accuracy. The mechanisms for the reward-induced decrease in discrimination accuracy (here observed primarily in males) are not fully understood. Rewards might temporarily lower the threshold for behavioral responding, resulting in stronger responding to weakly predictive cues [32]. Additionally, reward delivery might decrease discrimination accuracy by increasing the predictive value of the cue (and context) present during reward delivery. These direct cue→outcome (and context→outcome) associations compete with hierarchical associations for the control of behavior [33], potentially affecting responding on the next trial (e.g., a rewarded LIGHT: X+ trial might increase the value of cue X and promote responding during the following DARK: X-trial; the same logic could be applied to the influence of the context). Further investigation is required to determine the precise mechanism(s) by which reward disrupt discrimination performance in the context-dependent discrimination task.

Acute stress disrupts context-gated reward-seeking in male but not female rats

Stress affects several cognitive processes (e.g. selective attention, spatial navigation, working memory), often in a sex-dependent manner [25, 26, 34]. However the effect of stress on contextual modulation, and possible sex-differences in this effect, have not been formally tested. Therefore, we tested the effect of an acute stress (90-min restrain) on discrimination performance (expressed as discrimination ratio), in a subset of animals from all training groups (Ctx-dep O1: n = 11 F + 15M; Simple discrimination: n= 6F + 6M; No discrimination: n= 8F +8M; Ctx-dep O1/O2: n= 7F + 7M) (Fig. 4). With the exception of the No Discrimination group, acute stress reduced discrimination accuracy in all tasks (main Stress effect: Ps ≤ 0.004). However, only in the Ctx-dep. O1 task did we observe a significant Stress x Sex interaction F(1, 24) = 5.665; P = 0.026) as only males showed reduced discrimination performance following acute stress (Stress vs. Baseline; Males: T(14) = 5.165; P < 0.001; Females: T(10) = 1.290; P = 0.209). No main effect or interaction with Sex was observed in any other task (Ps ≥ 0.178; note however the smaller sample size in all these other groups). Therefore, although acute stress disrupted discrimination performance in all tasks, females trained in Ctx-dep. O1 appear to be less sensitive to the disrupting effect of stress in this task.

Acute stress disrupts context-gated reward seeking in male but not female rats

(A-B) Time in port (% cue time) during the four different trial types (LIGHT: X / DARK: X / LIGHT: Y / DARK: Y) at baseline and after acute stress for male (A) and female (B) rats trained in each discrimination task. (C) Discrimination ratio at baseline and after acute stress, for male and female rats trained in each discrimination task. (D) Stress-induced disruption in discrimination performance (stress – baseline) for male and female rats trained in each discrimination task (horizontal bar = median difference score). Error bars indicate ± s.e.m; * P < 0.05; *** P < 0.001, t-tests.

Female rats show higher orbitofrontal engagement during context-gated reward predictions

The hierarchical gating of associative predictions has been linked to activity in the orbitofrontal cortex (OFC) [28–30]. Therefore, for a subset of animals (Ctx-dep. O1: n= 11 F + 9M; Simple discrimination: n= 7F + 9M; No discrimination: n= 7F +7M; Ctx-dep. O1/O2: n= 5F + 5M), we quantified c-Fos-immunoreactive (cfos+) cells in the OFC (medial, ventral and lateral subregions) following regular behavior testing (no-stress condition). We also quantified cfos+ in two control regions, M1J and S1J, where we did not expect task-specific activation. Compared to baseline (homecage controls), behavioral testing increased cfos+ expression in all regions of interest, regardless of task (Fig. 5; homecage c-Fos+ counts are shown but not included in the following statistical analyses). A RM-ANOVA revealed a main effect of Region F(4, 192) = 85.037; P ≤ 0.001) and a significant Task x Sex x Region interaction (F(12, 192) = 1.9; P = 0.036). Follow-up planned contrast analysis compared males and females within each task and brain region. In the Ctx-dep. O1 task, females showed increased c-fos+ expression compared to males, in all OFC subregions (T(14) ≥ 2.387; P ≤ 0.032). No sex differences were found in any other task (Ps ≥ 0.255; note however the smaller sample size of the Ctx-dep. O1/O2 group). Moreover, no sex differences were found in the control regions M1J and S1J, regardless of task (Ps ≥ 0.139).

Female rats show higher orbitofrontal engagement during context-gated reward seeking.

(A) Experimental design. Brains were collected 2h after the start of the behavioral session. Brain sections were stained for c-Fos (activity marker) and NeuN (general neuronal marker). (B) Behavioral performance (discrimination ratio) ratio during the final session, before tissue collection for male and female rats trained in each task. (C-G) c-Fos+ density in the primary jaw motor cortex (C), primary jaw somatosensory cortex (D), medial orbitofrontal cortex (E), ventral orbitofrontal cortex (F), and lateral orbitofrontal cortex (G) for males and females trained in each discrimination task. Error bars indicate ± s.e.m. Gray bands indicate average c-fos+ density for animals who remained in their home cage prior to tissue collection (homecage controls; n=5). *: P < 0.05 Males vs. Females, t-tests.

Discussion

We investigated sex differences in contextual modulation of cue-evoked reward predictions and its neural correlates. Male and female rats were trained on a context-dependent discrimination task (Ctx-dep. O1) in which a contextual feature informs the validity of two reward cues (i.e. bidirectional occasion-setting). Overall, males acquired context-gated behavior faster than females. However, once established, contextual control over behavior was more robust in females; it was more stable within a session (less influence of prior reward) and more resistant to acute stress. This superior contextual control achieved by females was accompanied by elevated c-Fos expression in all OFC subregions (medial, ventral and lateral). Critically, no sex differences (behavioral or neurobiological) were observed among animals trained on simpler discrimination tasks that do not engage contextual gating (Simple Discrimination, No discrimination, Ctx-dep. O1/O2). These results add to the growing body of literature on sex differences in motivated behaviors and value-based decision-making [17, 24]. Most of these sex differences were observed in behaviors informed by binary associations between cues (or actions) and rewards. To our knowledge, this study is the first to demonstrate sex-differences in the context-informed hierarchical gating of associative predictions, a process that is largely independent from binary associations [3, 27].

For Ctx-dep. O1, the context (LIGHT OR DARK chamber) is an occasion-setter, i.e. a stimulus that modulates the response to a target cue, presumably by gating the retrieval of the appropriate cue→outcome association. However, our task differs from classic occasion-setting preparations in several ways. Historically, Pavlovian modulation has been studied in preparations in which the modulating stimuli (the occasion-setter) exerts unidirectional modulation, either potentiating (A:X+ / X-) or reducing (A:X- / X+) the response to the target cue. In such preparations, the occasion-setter might also establish a direct association (excitatory or inhibitory) with the outcome [1, 35, 36], thereby entangling the contributions of hierarchical modulation and direct stimuli-evoked predictions. To circumvent this issue, occasion-setting preparations commonly introduce a temporal gap (5-30s) between the termination of a discrete occasion-setting stimulus and its target cue. This approach has generally been successful in limiting direct associations between the occasion-setter and the outcome, instead promoting hierarchical modulation processes [37, 38, but see 39]. This temporal gap however introduces a working-memory requirement (animals must remember if the target cue was preceded by the occasion-setter) which complicates results interpretations.

Here, we used a bidirectional occasion-setting preparation with a single reward outcome (Ctx-dep. O1) as an alternative approach to investigate contextual modulation [31]. In this paradigm, a contextual feature (e.g., LIGHT) positively modulates one target association (LIGHT: X+ / DARK: X-) and negatively modulates another association (LIGHT: Y- / DARK: Y+). Critically, there are no direct informative associations between contextual feature and the single reward outcome (rewards are equally frequent in the LIGHT or DARK contexts), so accurate discrimination necessitates hierarchical modulation [27]. Moreover, the contextual feature precedes and overlaps with the target cue, thereby eliminating working-memory requirements. Thus, sex differences observed in Ctx-dep. O1, strongly indicate sex differences in hierarchical (context-informed) associative modulations. Alternatively, males and females might differ in terms of attentional resources [40]. Indeed, unlike Simple Discrimination or No discrimination tasks, Ctx-dep. O1 requires high multimodal attention (animals must simultaneously attend to the visual context and auditory cues). However, no sex differences were observed in context-dependent discrimination with two outcomes (Ctx-dep. O1/O2), a task that shares similar attentional requirements with Ctx-dep. O1, but does not necessarily engage hierarchical contextual modulation. Indeed, in Ctx-dep. O1/O2, both the visual context and the auditory target cue are relevant, however the use of differential outcomes presumably allows animals to solve the task via simple (nonhierarchical) summation processes. The possibility of a nonhierarchical solution to the discrimination problem might explain why almost all rats trained in the Ctx-dep. O1/O2 task learned the discrimination and at much faster rate compared to Ctx-dep. O1 (but see [31] for alternative explanation of the differential outcome effect). Together, these results indicate that the sex differences observed here are not attributable to simple associative, motivational, working-memory, or attentional processes, but are specific to the neurocomputational operations required for the hierarchical, contextual control of behavior. Note that there is no a priori reason to believe that the nature of these computations differ between sexes (no sex differences in the fundamental mechanisms of context-gated predictions). Rather male and female rats appear to differ slightly in the learning rate and magnitude/ robustness of these computations. Moreover it remains to be determined if the sex differences in context-gated reward seeking are specific to the sensory modalities employed here or if those sex differences extend to other sensory modalities.

Sex differences in context-dependent behaviors have previously been reported in fear conditioning preparations where females often display generalized fear responses that evade precise contextual control [41–44]. Elevated baseline anxiety and fear reactivity in females might contribute to this relative failure to regulate fear responses [45, 46, but see 47]. However, we show here that reduced contextual control in females (at least in the early stages of acquisition) is not limited to aversive behaviors but extends to appetitive, reward-seeking behaviors. This suggests a broader, domain-general, delay in the engagement of contextual, top-down regulation processes in female rodents. This is consistent with higher sign-tracking tendencies in female rats [48–52] —another form of appetitive behavior that reflects reduced top-down cognitive control (opposed to goal-tracking behavior) [53–55]. Note that the reduced top-down contextual control displayed by females in our study was limited to delayed acquisition. Provided sufficient training, females actually established more robust contextual control over cue-evoked reward seeking than males.

The biological basis for the delayed acquisition of contextual control in females is unknown. Although estrous cycle had no effect on the expression of context-dependent behavior in female discriminators, fluctuating sex hormones (and resulting neurobiological changes) could have contributed to the slower acquisition of contextual control in females. Indeed, learning the hierarchical contextual rules for reward prediction is a complex and lengthy process that probably requires distributed plasticity across several brain regions including the striatum, amygdala, hippocampus and neocortex [3, 56]. Estrous cycle affects synaptic physiology and plasticity at all these sites via changes in dopamine reuptake [57, 58], neuronal excitability [59–63], spine density [64, 65], gene expression [66, 67], or neurogenesis [68]. This dynamic neurobiological environment might constitute a challenge for females attempting to solve the Ctx-dep. O1 task, and might contribute to their delayed acquisition of contextual control. Interestingly, this dynamic neurobiological environment might also contribute to the more stable performance ultimately achieved by females. Indeed, in deep learning models, the injection of random “noise” in (artificial) neural networks during training is commonly used to build redundancy and increase network performance and stability [69, 70]. This ‘robustness through perturbations’ approach often comes at the expense of learning speed, similar to what we observed in females. Alternatively, or in addition, androgens (male sex hormones) –known to contribute to the male advantage on certain spatial tasks-might have contributed to males’ faster acquisition of contextual control in this task [71, 72]. The role of these hormonal factors, as well as potential non-gonadal sex differences, in the trade-off between learning speed and performance stability remains to be confirmed experimentally.

Regardless of the precise mechanism, our results indicate that, compared to male rats, female discriminators ultimately achieved more stable contextual control over cued reward-seeking; their context-dependent discrimination performance was less affected by trial history (recent rewards) or by acute stress. This relative resilience of females trained in the Ctx-dep. O1 task is consistent with studies showing that females, while more sensitive to the arousing and anxiogenic effects of stress, are more resilient to the disrupting effects of stress on executive functions and memory [22, 26]. Sex differences in regional sensitivity to CRF and CRF receptor signaling might contribute to these sex differences in behavioral responses to stress [73].

The behavioral sex differences observed in this study were accompanied by neural sex differences. Compared to males, females displayed increased OFC activation during Ctx-dep. O1. Importantly, no behavioral or neural sex differences were observed for any other task. Together, these results strongly implicate the OFC in the sex-biased expression of contextual modulation of reward-seeking. This is consistent with recent studies demonstrating a critical role of the OFC in hierarchical control over cue-evoked behaviors [28–30]. More generally, the OFC is proposed to encode task structure including context-dependent contingencies [74–76]. Accordingly, the OFC appears critical for decision-making when optimal choices require knowledge of the broader task structure (i.e. when stimuli cannot be taken at face value) [77, 78]. Although causality remains to be established, our data suggest that the increased orbitofrontal engagement displayed by females might contribute to their resilient contextual regulation of behavior.

The root causes of the behavioral and neural sex differences in context-gated reward prediction are largely unknown and not addressed in this study. In principle, all the sex differences reported here (speed of acquisition, robustness of performance, and OFC activation) might be independent and engage different mechanisms. Alternatively, some of these effects might be causally related. Our result show that, for a comparable level of context-dependent discrimination, females show an increased engagement of the OFC compared to males (higher number of c-Fos immunoreactive neurons). This suggests a denser (less sparse) coding of context-dependent contingencies in females’ OFC. Denser coding is generally associated with slower learning but superior fault-tolerance (i.e. better resistance to noise or disturbance) [79], exactly what we observed in females, suggesting a potential unifying mechanism for the sex differences reported here.

The contextual regulation of behavior is at the heart of cognitive control [12, 80]. Deficits in this process are integral to the pathophysiology of numerous neuropsychiatric disorders [4]. Although extrapolation of animal data to humans requires caution, our findings of a superior OFC engagement by female rats during context-informed hierarchical control of reward seeking is consistent with human [f]MRI studies showing larger OFC grey matter volume in females [81] and superior OFC activation in females during the regulation of emotional responses [82, 83]. The behavioral and neural sex differences observed here might offer insight into the biological sex differences observed in certain disorders. For instance, the initial delay in establishing contextual control observed here in females might contribute to the rapid progression of SUD or PTSD observed in women following initial drug use or trauma [14, 15, 84, 85]. Conversely, the more stable contextual control ultimately achieved by females after extended training observed here might contribute to the better prognosis for women following behavioral and cognitive treatments for these disorders [85–87], and might be due, in part, to increased OFC engagement.

Methods

Subjects

Subjects were male and female Long Evans rats (Charles Rivers), 8-10 weeks at arrival. Rats were housed in same-sex pairs in a temperature-controlled vivarium (22 °C) with a 12-hr light/dark cycle; behavioral experiments were conducted during the light cycle. Rats were given free access to water in their home cage for the duration of the experiments. After a week of acclimation to the colony room, rats were mildly food-restricted to maintain ∼95% of age-matched free-feeding weights. All experimental procedures were conducted in accordance with UCSB Institutional Animal Care and Use Committees and the US National Institute of Health guidelines.

Apparatus

Behavioral training was conducted in 12 identical conditioning chambers enclosed in individual sound- and light-attenuating cubicles (Med Associates, St. Albans, VT). Six chambers were dedicated to male subjects and six chambers were dedicated to female subjects; animals of both sexes were trained simultaneously. A recessed reward delivery port was located at the center of the front panel and was connected to a pellet dispenser and a syringe pump for liquid reward delivery (syringe pump located outside of the sound-attenuating cubicle). Subjects’ presence in the reward port was detected by the interruption of an infrared beam. Two ceiling-facing chamber lights located on the front and back internal walls of the cubicle provided diffuse chamber illumination and were used to manipulate the background visual context. Auditory stimuli were delivered via a white noise speaker and a clicker (76dB each), located in the back and front panel, respectively. A fan mounted on the cubicle provided ventilation and low background noise. A computer equipped with Med-PC software (Med Associates) controlled all experimental programs and data collection.

Rewards

Food rewards consisted of a 45mg banana-flavored food pellet (BioServe Dustless Precision Pellets, F0059 formula) or a 0.17mL bolus of sweetened chocolate milk (2 parts chocolate Nesquik® + 1 part water + 5% sucrose w/v) delivered over 3s. These rewards are highly discriminable but equicaloric and were equally preferred by rats in preliminary studies. The identity of the rewarding outcomes was counterbalanced within behavioral tasks and within sexes.

Discrimination Training

Rats were randomly assigned to one of four Pavlovian discrimination conditions: Context dependent discrimination with single outcome (Ctx-dep. O1, n = 30F + 32M), Context dependent discrimination with dual outcome (Ctx-dep. O1/O2, n = 14F + 14M), Simple discrimination (Simple, n = 12F + 22M), or No Discrimination (NoDiscr., n = 8F + 8M). Animals were individually placed in the conditioning chambers for 80-min training sessions 5-6 days/week. During each session, the visual background –the local context– alternated every two minutes between flashing houselights (0.25s on, 0.25s off; LIGHT context) or darkness (DARK context). Within each context phase, two 10-s auditory cues, X or Y (white noise or clicker, counterbalanced), were presented individually, in two separate trials, and were potentially followed by reward. The presentation of each cue within a context phase was selected semi-randomly, so that the identity of the first cue had no incidence on the identity of the second cue (i.e. the same cue could repeat twice within a context phase, so the outcome of the first trial did not inform the outcome of the second trial). Each session consisted of 40 context cycles and 80 cue presentations (i.e. 80 trials) with an average intertrial of 50s ± 20s (rectangular distribution). The reward contingencies in each task are outlined in Figure 1B. During the acquisition period, unless specified otherwise (cf. Ctx-dep. O1), each behavioral training session consisted of an equal proportion of each trial type (20* LIGHT: X; 20* DARK: X; 20* LIGHT: Y; 20* DARK: Y).

Context dependent discrimination with single outcome (Ctx-dep. O1)

In this task, the reward-predictive validity of each cue was informed by the context, so that cue X was rewarded only in LIGHT context and cue Y was rewarded only in the DARK context. All rewarded trials resulted in the same reward outcome (O1). Considered individually, both cues and both contexts have equal probabilities of resulting in the same outcome, therefore accurate performance in this task cannot be understood in terms of summed associative strength between cues and contexts. Instead, accurate performance in this task critically relies on contextual modulation (i.e. occasion-setting). During the first 35 training sessions, rats experienced an equal number of rewarded and unrewarded trials (20 trials for each context/cue combination per session). On the 36th session and for the remainder of the experiment, the relative proportion of rewarded trials was reduced (10* LIGHT: X+; 30* DARK: X-; 30* LIGHT: Y-; 10* DARK: Y+) in an effort to promote discrimination. After 72 training sessions, animals that failed to reach a discrimination criterion were excluded from the rest of the study (discrimination criterion: time in port during rewarded cue ≥ 1.5* time in port during nonrewarded cue, for both cues, for a minimum of 4 out of 5 consecutive sessions).

Simple Discrimination

Cue X was always rewarded and cue Y was never rewarded, regardless of the context. Presumably, this training protocol promotes accurate reward prediction via simple associative learning (context-independent predictions).

No Discrimination

Both cues X and Y were probabilistically rewarded in both context (the background context provides no disambiguating information). During the acquisition period, the probability of reinforcement was 50% for each cue/ context combination. Presumably, this training protocol promotes reward prediction via simple associative learning (context-independent predictions) although the reward cues retain some ambiguity.

Context dependent discrimination with dual outcome (Ctx-dep. O1/O2)

This task followed similar contingencies as Ctx-dep. O1, with the exception that the two rewarded trial types (LIGHT: X and DARK: Y) resulted in two different food outcomes (O1 and O2, respectively). This has profound implications for the learning processes engaged in the task. The delivery of different outcomes in different contexts allows for distinct context-outcome associations (LIGHT→O1 and DARK→O2), as well as distinct cue-outcome associations (X→O1 and Y→O2). These different associative structures could potentially facilitate discrimination performance via simple summation process.

Assessment of Estrous Cycle

Estrous cycle phases were monitored via vaginal cytology for a subset of females (n = 10) over 15 consecutive days (sessions 66-80). Immediately after the behavioral session, the tip of a saline-moistened cotton swab was inserted into the vagina and rotated to dislodge cells from the vaginal wall. Swabs were immediately rolled onto a glass slide, and the samples were preserved with a spray fixative (M-FixTM, MilliporeSigma) without allowing the cells to dry. Slides were then stained with a modified Papanicolau staining procedure as follows: 50% ethyl alcohol, 3 min; tap water, 10 dips (×2); Gill’s hematoxylin 1, 6 min; tap water, 10 dips (×2); Scott’s water, 4 min; tap water, 10 dips (×2); 95% ethyl alcohol, 10 dips (×2); modified orange-greenish 6 (OG-6), 1 min; 95% ethyl alcohol, 10 dips; 95% ethyl alcohol 8 dips; 95% ethyl alcohol 6 dips; modified eosin azure 36 (EA-36), 20 min; 95% ethyl alcohol, 40 dips; 95% ethyl alcohol, 30 dips; 95% ethyl alcohol, 20 dips; 100% ethyl alcohol, 10 dips (×2); xylene, 10 dips (×2); coverslip immediately. Samples were observed with light microscopy (4x). Four criteria were used to determine the specific estrus cycle: 1. cell density, 2. percentage of nucleated epithelial cells, 3. cornified epithelial cells, and 4. Percentage of leukocytes [88–90]. Each sample was independently rated by two trained observers (S.P. and A.M.), blind to the behavioral data. Inter-rater agreement was high (88.2%). Only behavioral sessions for which both raters agreed on the estrous phase were selected for analysis. As a result, 2-4 behavioral sessions were considered for each subject at each phase of the estrous cycle. Behavioral data on these sessions were averaged to express a single data point for each rat at each stage of the estrous cycle. Females that did not show a regular cycle were excluded from the analysis (n = 2).

Acute Restraint Stress

Before testing the effect of acute stress on discrimination performance, the relative proportion of rewarded trials was adjusted in all groups to match the proportion of rewarded trials in the context-dependent discrimination with a single outcome (Ctx-dep. O1). As a result, rewarded cues became less frequent and represented only 1/4 of the trials. For all groups, these rewarded trials were equally distributed across both contexts. After 5-6 sessions of adjustment to these new conditions, we tested the effect of an acute stress on discrimination performance in a subset of animals from all training groups (Ctx-dep. O1: M n=15, F n=11; Simple discrimination: M n=6, F n=6; No discrimination: M n=8, F n=8; Ctx-dep. O1/O2: M n=7, F n=7). Acute stress was induced by restraining rats in standard Plexiglas rat restraint tubes (Plas Labs, IncTM) for 90 min, in a novel brightly lit room. To account for the size differences between male and females, two different sized restrainers were used (65 and 50mm internal diameter for male and female restrainers respectively). After placing the rats in the tube, the length of the restrainer was adjusted to immobilize the rat without causing pain or interfering with animals’ breathing. After 90 min of restraint, the rats were returned to their homecage and left undisturbed for 5 min before being placed in the conditioning chambers for a normal behavioral session.

cFos Immunofluorescence

Following stress test, animal were retrained for 5-6 sessions, under standard conditions (no stress; equal number of rewards for all training groups; rewarded trials = 1/4 of all trials). After their last behavioral session, animals remained in the conditioning chambers for 40 min (120 min total); they were then deeply anesthetized with a lethal dose of Euthasol and perfused transcardially with ice-cold 1x PBS followed by ice-cold 4% PFA in PB. Brains were extracted, post-fixed (4% PFA overnight) and cryoprotected (30% sucrose + 0.1% sodium azide, 1x PBS) until processed for immunostaining. Brains were frozen and sectioned into 40μm with a sliding cryostat (Leica CM1860). Free floating coronal sections were washed for 10 min in PBS, followed by a 2-h incubation in a blocking solution (5% NGS, 0.2% Triton X-100 in 1x PBS) at room temperature. Sections were then incubated with rabbit anti-cFos (1:1000, Cell Signaling, CAT#: 2250) and mouse anti-NeuN (neuron-specific protein; 1:2000 or 1:3000, Novus, CAT#: NBP1-92693) primary antibodies, in the blocking solution, for 20 h a 4°C. Following primary antibody incubation, sections were washed in PBS 3 times for 10 min and then incubated with anti-rabbit Alexa Fluor® 647-conjugated secondary antibody (far-red) (1:250, Jackson ImmunoResearch, CAT#: 111-605-144) and anti-mouse Alexa Fluor® 488-conjugated secondary antibody (green) (1:500, Jackson ImmunoResearch, CAT#: 115-545-003) for 90 min at room temperature. Sections were then washed in PBS three times for 10 min, mounted on slides and immediately coverslipped with a hardset mounting medium (Vector laboratories CAT#: H-1700). Images were acquired at a 10X magnification, using a Keyence BZ-X fluorescence microscope.

Brain regions of interest for c-Fos immunoreactive neurons quantification included the medial, ventral, and lateral orbitofrontal cortex (mOFC, vOFC, and lOFC). Additionally, two control regions were included, for which we did not expect task-specific activation: the jaw primary motor (M1J) and somatosensory cortex (S1J). These regions of interests were manually isolated from the original images, using only the green / NeuN channel (the far-red / c-fos channel was temporarily disabled for blinding purposes). A custom ImageJ script (National Institute of Health) was then used to count c-Fos positive neurons in the selected regions and normalize this count to the surface area of the selected region (c-Fos positive neurons / mm2). For each brain region, c-Fos positive neurons were counted bilaterally from 2-4 sections per animal and a mean count was computed for each animal.

Statistical analysis

Data were collected over three replication cohorts, each composed of approximately equal number of male and female rats. Cue-evoked reward seeking was quantified primarily as the percentage of time a rat spent in the food cup during the last 5 seconds of cue presentation, when anticipatory food cup behavior is more reliably observed and less contaminated by orienting behaviors [91–93]. The percentage of time in port during the 5s epoch prior cue onset is shown but not analyzed. Discrimination accuracy was expressed as a discrimination ratio, defined as the time in port during rewarded trials divided by the total time in port during all trials. A ratio of 0.5 indicates equal responding during rewarded and non-rewarded trials (i.e. no discrimination) whereas a ratio of 1 indicates responding exclusively during rewarded trials (i.e. perfect discrimination). Time in port and discrimination ratio during acquisition were aggregated in 4-session bins. These longitudinal repeated measures were analyzed with Linear Mixed-Models (LMM) with a Restricted Maximum Likelihood (REML) estimation method. All covariance structures were explored and the best fitting model was determined by the lower Akaike information criterion score [94, 95]. The selection of a covariance structure influenced the degrees of freedom of the model, possibly resulting in non-integer values. The remaining of the analyses consisted generally of mixed-models repeated measures (RM) ANOVAs with Sex and/or Task as a between-subject factor, and Trial history, Stress, or Brain region as within-subject factors. The Geisser-Greenhouse correction was applied when data lacked sphericity, possibly resulting in non-integer degrees of freedom. Post-hoc or planned contrast pairwise comparisons were carried with Bonferroni-corrected t-tests. When appropriate, nonparametric tests were conducted and consisted of the Kruskal-Wallis H test followed by post-hoc Bonferroni-corrected Dunn’s pairwise comparisons. All tests were two-tailed, significance was assessed against a type I error rate of 0.05. Statistical analyses were conducted with SPSS Statistics package (version 28.0.0.0; IBM SPSS). Graphs were generated using GraphPad Prism 9.

eLife Assessment

This valuable manuscript reveals sex differences in bi-conditioning Pavlovian learning and conditional behavior. Males learn hierarchical context-cue-outcome associations more quickly, but females show more stable and robust task performance. These sex differences are related to cellular activation in the orbitofrontal cortex. Although the evidence for the claims is convincing, the claim of sex differences in context-dependent discrimination behaviour is overstated in places. Nevertheless, the results will be of interest to many behavioural neuroscientists, particularly those who investigate sex-specific behaviours.

Reviewer #1 (Public Review):

Summary:

Peterson et al., present a series of experiments in which the Pavlovian performance (i.e. time spent at a food cup/port) of male and female rats is assessed in various tasks in which context/cue/outcome relationships are altered. The authors find no sex differences in context-irrelevant tasks, and no such differences in tasks in which the context signals that different cues will earn different outcomes. They do find sex differences, however, when a single outcome is given and context cues must be used to ascertain which cue will be rewarded with that outcome (Ctx-dep O1 task). Specifically, they find that males acquired the task faster, but that once acquired, performance of the task was more resilient in female rats against exposures to a stressor. Finally, they show that these sex differences are reflected in differential rates of c-fos expression in all three subregions of rat OFC, medial, lateral and ventral, in the sense that it is higher in females than males, and only in the animals subject to the Ctx-dep O1 task in which sex differences were observed.

Strengths:

• Well written

• Experiments elegantly designed

• Robust statistics

• Behaviour is the main feature of this manuscript, rather than any flashy techniques or fashionable lab methodologies, and luckily the behaviour is done really well.

• For the most part I think the conclusions were well supported, although I do have some slightly different interpretations to the authors in places.

Weaknesses:

The authors have done an excellent job of addressing all previous weaknesses. I have no further comments.

Reviewer #2 (Public Review):

Summary:

A bidirectional occasion-setting design is used to examine sex differences in the contextual modulation of reward-related behaviour. It is shown that females are slower to acquire contextual control over cue-evoked reward seeking. However, once established, the contextual control over behaviour was more robust in female rats (i.e., less within-session variability and greater resistance to stress) and this was also associated with increased OFC activation.

Strengths:

The authors use sophisticated behavioural paradigms to study the hierarchical contextual modulation of behaviour. The behavioural controls are particularly impressive and do, to some extent, support the specificity of the conclusions. The analyses of the behavioural data are also elegant, thoughtful, and rigorous.

Comments on revised version:

In this revised version the authors have addressed the major weaknesses that I identified in my previous review.

Reviewer #3 (Public Review):

Summary:

This manuscript reports an experiment that compared groups of rats acquisition and performance of a Pavlovian bi-conditional discrimination, in which the presence of one cue, A, signals that the presentation of one CS, X, will be followed by a reinforcer and a second CS, Y, will be nonreinforced. Periods of cue A alternated with periods of cue B, which signaled the opposite relationship, cue X is nonreinforced and cue Y is reinforced. This is a conditional discrimination problem in which the rats learned to approach the food cup in the presence of each CS conditional on the presence of the third background cue. The comparison groups consisted of the same conditional discrimination with the exception that each CS was paired with a different reinforcer. This makes the problem easier to solve as the background is now priming a differential outcome. A third group received simple discrimination training of X reinforced and Y nonreinforced in cues A and B, and the final group were trained with X and Y reinforced on half the trials (no discrimination). The results were clear that the latter two discrimination learning procedures resulted in rapid learning in comparison to the first. Rats required about 3 times as many 4-session blocks to acquire the bi-conditional discrimination than the other two discrimination groups. Within the biconditional discrimination group, female and male rats spent the same amount of time in the food cup during the rewarded CS, but females spent more time in the food cup during CS- than males. The authors interpret this as a deficit in discrimination performance in females on this task and use a measure that exaggerates the difference in CS+ and CS_ responding (a discrimination ratio) to support their point. When tested after acute restraint stress, the male rats spent less time in the food cup during the reinforced CS in comparison to the female rats, but did not lose discrimination performance entirely. The was also some evidence of more fos positive cells in the orbitofrontal cortex in females. Overall, I think the authors were successful in documenting performance on the biconditional discrimination task, showing that it is more difficult to perform than other discriminations is valuable and consistent with the proposal that accurate performance requires encoding of conditional information (which the authors refer to as "context"). There is evidence that female rats spend more time in the food cup during CS-, but this I hesitate to agree that this is an important sex difference. There is no cost to spending more time in the food cup during CS- and they spend much less time there than during CS+. Males and females also did not differ in their CS+ responding, suggesting similar levels of learning, A number of factors could contribute to more food cup time in CS-, such as smaller body size and more locomotor activity. The number of food cup entries during CS+ and CS- was not reported here. Nevertheless, I think the manuscript will make a useful contribution to the field and hopefully lead readers to follow up on these types of tasks. One area for development would be to test the associative properties of the cues controlling the conditional discrimination, can they be shown to have the properties of Pavlovian occasion setting stimuli? Such work would strengthen the justification/rationale for using the term "context" and "occasion setter" to refer to these stimuli in this task in the way the authors do in this paper.

Strengths:

Nicely designed and conducted experiment.

Documents performance difference by sex.

Weaknesses:

Overstatement of sex differences.

Inconsistent, confusing, and possibly misleading use of terms to describe/imply the underlying processes contributing to performance.

Author response:

The following is the authors’ response to the previous reviews.

Reviewer 2 (Public Review):

Stress response in males versus females: The authors argue that the contextual control over behaviour was more robust in female rats as females show less within session variability and greater resistance to stress. What evidence is there that the restraint stress procedure caused a similar stress response in both sexes? That is, was the stress induction equally effective in males and females?

The restraint protocol used in this study is a well-established stressor in rodents, known to produce robust behavioral and physiological effects (HPA axis activation), in both sexes. Although not measured in this study, the ACTH and cortisol responses are actually greater in females during restraint. To the extent that “stress induction” is interpreted as “HPA axis activation”, this strongly suggests that the stress induction in males and females was at least comparable, if not greater in females.

We have added a few sentences (in the Result and Method section) to highlight this important point. We thank the reviewer for bringing this up.

Minor corrections:

(1) Please verify that the in-text reference to the figures is correct. I noticed a few mistakes, for example:

- Line 120 (pdf) refers to Fig. 1 C-D but should refer to D only.

- Line 312 (pdf) refers to Fig 1D for discrimination ratios but these are shown in Fig 1E

- No reference in text to 2A

Thank you for bringing this to our attention. We have fixed the in-text references to the figures.

(2) In the results it states that the homecage c-Fos+ counts are shown in Figure 5 but I couldn't see these?

The homecage c-Fos+ counts were initially shown as a pale gray band in the background of the main histograms. Because those counts are very low, it was hard to dissociate this gray band from the black horizontal axis. We have replaced the gray band with a more vivid blue line that is now in the foreground of the histograms. Moreover, we added a note in the figure legend to bring readers’ attention to this homecage count line, close to floor level.

(3) Line 306: It is stated that "the use of differential outcomes presumably allows animals to solve the task via simple (nonhierarchical) summation processes". I don't understand the use of "summation" here, isn't it simply that the rats are relying on direct context-outcome and/or cue-outcome associations?

That’s right. These rats might be relying on direct context-outcome and cue-outcome associations and adding (or summing up) the converging expectations. We have added a few words in the text to clarify what we mean by summation (i.e. the addition of converging cue-evoked + context-evoked predictions).

Version 3

Abstract

Animals, including humans, rely on contextual information to interpret ambiguous stimuli. Impaired context processing is a hallmark of several neuropsychiatric disorders, including schizophrenia, autism spectrum disorders, post-traumatic stress disorder, and addiction. While sex differences in the prevalence and manifestations of these disorders are well established, potential sex differences in context processing remain uncertain. Here we examined sex differences in the contextual control over cue-evoked reward seeking and its neural correlates, in rats. Male and female rats were trained in a bidirectional occasion-setting preparation in which the validity of two auditory reward-predictive cues was informed by the presence, or absence, of a visual contextual feature (LIGHT: X+ / DARK: X− / LIGHT: Y− / DARK: Y+). Females were significantly slower to acquire contextual control over cue-evoked reward seeking. However, once established, the contextual control over behavior was more robust in female rats; it showed less within-session variability (less influence of prior reward) and greater resistance to acute stress. This superior contextual control achieved by females was accompanied by an increased activation of the orbitofrontal cortex compared to males. Critically, these behavioral and neural sex differences were specific to the contextual modulation process and not observed in simple, context-independent, reward prediction tasks. These results indicate a sex-biased trade-off between the speed of acquisition and the robustness of performance in the contextual modulation of cued reward seeking. The different distribution of sexes along the fast learning ↔ steady performance continuum might reflect different levels of engagement of the orbitofrontal cortex, and might have implications for our understanding of sex differences in psychiatric disorders.

Introduction

Pavlovian learning is often conceived as the formation of a binary association between two events, where event #1 (the stimulus) evokes the anticipation of event #2 (the outcome). Under this naive definition, the background circumstances (the context) in which those events occur play little role. However, this simple form of learning is often augmented by higher level cognitive systems that allow animals (including humans) to learn and retrieve context specific associations (e.g., the word “shot” might evoke different representations in a bar vs. a vaccine clinic) [1–3]. This process is referred to as “contextual-gating” (the context acting as a “gatekeeper” in the selection of the appropriate associative memory).

Deficits in contextual gating are observed in several neuropsychiatric disorders including schizophrenia, autism spectrum disorder (ASD), post-traumatic stress disorders (PTSD), and substance use disorders (SUD) [4–12], and may contribute to the intrusive thoughts and maladaptive (situation-inappropriate) behaviors that characterize these disorders. Importantly, the prevalence and symptomatology of these disorders vary between sexes. For example, the incidence of schizophrenia and ASD is higher in men [13] while women are more at risk of developing SUD or PTSD following initial drug use or trauma [14, 15]. This disparity might be due, in part, to subtle sex differences in executive strategies [16–19]. Stress, a contributing factor in all aforementioned disorders, can further exacerbate these sex differences, resulting in sex-specific cognitive vulnerabilities [18–23]. Sex differences in executive functions and sensitivity to stress have been reproduced in animal models using behavioral assays of attention, impulsivity, spatial navigation, and working memory [17, 24–26]. Whether these sex differences extend to the contextual gating of associative predictions remains unknown.

In animal models, the contextual gating of associative predictions can be studied in occasion-setting preparations. In positive occasion-setting, animals learn that a “target” cue (X) results in a reward outcome (+) only when that cue is accompanied by a specific “feature” such as a specific context (A); the same cue presented in absence of this contextual feature remains without consequence (A:X+ / X-). Conversely, in negative occasion-setting, a target cue results in reward, except when accompanied by a specific feature (A:X- / X+). In both preparations, the contextual feature is thought to hierarchically modulate (positively or negatively) the association between the target cue and the outcome [3, 27]. Strong evidence for this hierarchical modulation is the fact that a contextual feature can simultaneously function as a positive modulator for one target cue and as a negative modulator for another target cue (A:X+ / X- / A:Y- / Y+). In such bidirectional occasion-setting procedures all cues and contexts have equal probabilities of reward, therefore, accurate reward predictions cannot be achieved via simple associative summation but instead require the context-informed hierarchical gating of associative predictions (i.e. context-gated associative predictions).

Compared to simple forms of associative learning, the neural bases of context-gated associative predictions (or occasion-setting) remain severely understudied. Although recent studies have implicated the orbitofrontal cortex (OFC) in occasion-setting [28–30], most of the behavioral and neurobiological occasion- setting studies have exclusively used male subjects and/or did not explicitly address sex as a biological variable.

Therefore, the purpose of this study was two-fold: 1) investigate sex differences in the (bidirectional) contextual gating of reward prediction and its sensitivity to stress, and 2) investigate sex differences in OFC engagement during the contextual gating of reward predictions. We found that female rats were slower than males to acquire context-gated reward-seeking. However, once established, context-gated performance was more stable in females; it showed less variability within a session and was more resistant to acute stress. Moreover, we showed that this superior contextual control achieved by females was accompanied by an increased activation of the orbitofrontal cortex.

Results

Male and female rats were randomly assigned to one of four Pavlovian discrimination tasks outlined in Fig. 1A-C. In these tasks [adapted from 31], all rats were exposed to two auditory cues (X and Y, 10s) presented within two distinct, alternating, visual contexts (LIGHT and DARK; 2min). All rats received similar amount of reward, and produced a similar behavioral response (magazine approach). However, the different reward contingencies featured in these tasks presumably promoted the emergence of different associative architectures for reward prediction (Fig. 1B).

Sex differences in the acquisition of context-gated but not context-independent reward predictions.

(A) Apparatus and general task design. During each session, the local context alternates every 2min between a Ctx. LIGHT (houselight on) and Ctx. DARK (houselight off). Embedded in these context cycles, brief auditory stimuli (X or Y; 10s) are presented and potentially followed by reward (chocolate milk or banana pellets). (B) Reward contingencies and corresponding associative structures for each training condition. Ctx-dep. O1: the local context informs (hierarchically “gates”) the predictive status of the ambiguous reward cues X and Y. Simple discrimination: reward cues are not ambiguous (only X is rewarded) and the local context is irrelevant. No Discrimination: reward cues X and Y are potentially rewarded and the local context is irrelevant. Ctx- dep. O1/O2: Context-dependent discrimination with differential outcomes; in this case, the local context is relevant but may contribute to context-sensitive performance via simple (non-hierarchical) associations (summation of cue- and context- evoked predictions). (C) Experimental timeline. (D) Time in port (% cue time) during rewarded and non-rewarded trials across acquisition for male and female rats trained in each task. (E) Discrimination ratio across acquisition for male and female rats trained in each task. Error bars and error bands indicate ± s.e.m. *: P < 0.05, t-tests

In the Ctx-dep. O1 task, cue X was rewarded only in the LIGHT context and cue Y was rewarded only in DARK context; all rewarded trials resulted in the same outcome (O1). Accurate reward prediction in this task critically relies on the context-informed hierarchical modulation associative predictions (i.e occasion- setting). In the Simple discrimination task, cue X was always rewarded and cue Y was never rewarded, regardless of context. In the No discrimination task both cues X and Y were probabilistically rewarded in both contexts. Finally, the Ctx-dep. O1/O2 task featured similar contingencies as Ctx-dep. O1, with the exception that the two rewarded trials (LIGHT: X and DARK: Y) resulted in two different outcomes (O1 and O2, respectively). In this case, not only might the visual context promote the hierarchical modulation of associative predictions (as in Ctx-dep. O1), but the context might also contribute to context- sensitive performance via simple (non-hierarchical) associations. Indeed, the use of differential outcomes allows for distinct context-outcome associations (LIGHT→O1 and DARK→O2), as well as distinct cue-outcome associations (X→O1 and Y→O2). On a given trial, if the cue and context tend to activate the same outcome representation, the converging cue-evoked and context-evoked excitations can add up. As a result, context-sensitive performance could potentially be achieved via simple (nonhierarchical) summation process. Therefore, while rats trained in Ctx-dep. O1/O2 might engage a combination of associative processes to achieve context-sensitive behavior (including hierarchical and nonhierarchical associations), only rats trained in the Ctx-dep. O1 task critically and unambiguously rely on hierarchical associations to achieve context-sensitive behavior.

Male rats acquire context-regulation over reward- seeking faster than females

Fig. 1D displays the course of acquisition of discriminated Pavlovian approach in all four tasks (Ctx- dep. O1 n = 30F + 32M; Simple discrimination n = 12F + 22M; No discrimination n = 8F + 8M; Ctx-dep. O1/O2 n = 14F + 14M). In the context-dependent discrimination with a single outcome (Ctx dep. O1), Linear Mixed Models (LMM) analysis revealed a significant Trial x Session interaction, (F(17, 181.488) = 73.343, P < 0.001), as rats gradually learned to respond more during rewarded than non-rewarded trials. Critically, the analysis also revealed a significant Sex x Trial x Session interaction (F(17, 181.488) = 4.683, P < 0.001). Post-hoc comparisons revealed that males and females displayed similar levels of responding to rewarded trials (Ps > 0.260), but females maintained higher responding during non-rewarded trials, causing both sexes to differ in this measure from session 40 onwards (Ps < 0.047).

To facilitate comparisons between groups, we calculated for each animal a discrimination ratio, defined as the time in port during rewarded trials divided by the total time in port during all trials (Fig 1E). With the exception of animals trained in the No Discrimination task (for which discrimination between rewarded and non-rewarded trials was impossible by design), all training groups demonstrated a gradual increase in discrimination ratio (main Session effect: No Discr. P = 0.477; all other groups Ps < 0.001). Importantly, only in Ctx-dep. O1 did we detect a main effect of Sex (F(1, 69.876) = 15.185, P < 0.001) and a Sex x Session interaction (F(17, 365.569) = 3.054, P < 0.001). Indeed, acquisition of discriminated performance was faster in males, causing significant sex differences from session 28 onwards (Ps < 0.014). No main effect or interaction with Sex was detected for any other training group (Ps > 0.305).

Fig. 2A summarizes the performance of each group (time in port during the four trial types, LIGHT: X / DARK: X / LIGHT: Y / DARK: Y) by the end of the acquisition period (i.e. after 72 sessions for Ctx-dep. O1, or 32 session for all other groups). We next calculated the proportion of male and female rats that successfully acquired discriminated responding during the training period (Fig. 2B-C). The criterion for successful discrimination was defined as a time in port during rewarded trials that exceeded the time in port during non-rewarded trials by at least 50% (Time rew ≥ 1.5* Time nonrew, which corresponds to a discrimination ratio ≥ 0.6) for a minimum of 4 out of 5 consecutive sessions. Rats in the Ctx-dep. discrimination groups had to reach this criterion for both cues to be considered “discriminators” (LIGHT:X ≥ 1.5* DARK:X and DARK:Y ≥ 1.5* LIGHT:Y). A significant sex difference in the proportion of discriminators was observed in the Ctx- dep. O1 group (χ2 = 5.35; P = 0.021), as a greater proportion of males reached criterion (Fig. 2 B-C). No sex differences were observed in the proportion of discriminators in any other tasks.

Individual differences in the acquisition of discriminated Pavlovian approach.

(A) Time in port (% cue time) during the four different trial types (LIGHT: X / DARK: X / LIGHT: Y / DARK: Y) at the end of the training period (last 3 days average) for male and female rats trained in each discrimination task. (B) Discrimination ratio at the end of the training period (last 3 days average) for male and female rats trained in each discrimination task. Dark symbol: discriminator; Grey symbol: non-discriminator. Horizontal bar = group median; Dashed horizontal line represents the indifference level (no discrimination). (C) Proportion of male and female rats classified as discriminator or non-discriminator in each task. D Speed of acquisition (sessions to criterion) for discriminator rats, in each task. Horizontal bar = group median *: P < 0.05 Males vs. Females, t-tests; §: P < 0.05 chi-square test.

We then analyzed the number of sessions required for rats to reach the discrimination criterion (Fig. 2D). Rats that failed to reach the discrimination criterion were excluded from this analysis. A Kruskal-Wallis one-way ANOVA revealed a significant effect of the behavioral task (H (2) = 87.361, P < 0.001) with different sessions- to-criterion for all the discrimination tasks (Ctx-dep. O1 > Ctx-dep. O1/O2 > Simple; Ps < 0.001, Dunn’s tests). Note how the introduction of differential outcomes profoundly facilitated the acquisition of context- dependent discrimination (sessions to criterion: 47.8 ± 1.7 s.e.m. for Ctx-dep. O1 vs. 17.26 ± 1.6 s.e.m. for Ctx-dep. O1/O2). This is consistent with a prior study showing a similar differential outcome effect in this task [31].

Sex differences in the speed of acquisition (i.e. sessions-to-criterion) were analyzed separately within each task. In Ctx-dep. O1, males were significantly faster to reach the discrimination criterion (T(38) = 2.502, P = 0.017). No sex differences in the speed of acquisition were observed in the other tasks (Ps ≥ 0.175).

These results indicate that female rats were generally slower to acquire contextual control over cue-evoked reward-seeking. This sex difference in the development of discriminated performance was observed only in Ctx- dep. O1, the only task that unambiguously engages contextual modulation of reward predictions. Note that estrous cycle did not affect performance in female discriminators (Fig. S1). Rats that failed to reach the discrimination criterion at the end of the acquisition period were excluded from further experimental procedures (with the exception of rats trained in the No Discrimination group).

Trial history influences context-gated reward-seeking in male but not female rats

The delivery of reward, while critical for learning, can also interfere with the expression of discrimination behavior [32]. Therefore, we investigated to which extent within-session reward history (i.e. prior rewarded or non-rewarded trials) influenced performance in the Ctx-dep. O1 task, in male (n=25) and female (n=15) discriminators (Fig. 3). Performance was analyzed over 10 consecutive sessions (sessions 71-80; 80 trials/ session; 800 trials total), a period during which discrimination performance was globally stable. The first trial of a session was excluded as it was not preceded by any trial. For the remaining 790 trials (10 sessions x 79 trials), we determined if the trial was preceded by a rewarded or non-rewarded trial. This large sample size is required to ensure that enough data was collected for each possible trial history scenario. Responding during non-rewarded and rewarded trials was analyzed separately.

Recent rewards disrupt contextual control of reward seeking in male, but not female, rats.

(A) Time in port (% cue time) during non-rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR) for male and female rats trained in Ctx-dep. O1 task. Inset: increase in responding caused by prior reward for male and female rats (horizontal bar = median difference score). (B) Time in port during non-rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. (C) Response probability on non-rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: increase in the probability of responding caused by prior reward for male and female rats (horizontal bar = median difference score). (D) Response probability on non-rewarded trials as a function of prior consecutive non- rewarded or rewarded trials. (E) Time in port (% cue time) during rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: difference in responding caused by prior reward for male and female rats. (F) Time in port during rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. (G) Response probability on rewarded trials as a function of prior trial (rewarded: prior R, or non-rewarded: prior NR). Inset: difference in the probability of responding caused by prior reward for male and female rats. (H) Response probability on rewarded trials as a function of prior consecutive non-rewarded or rewarded trials. Error bars indicate ± s.e.m. * P < 0.05 Males vs. Females; # P < 0.05; # # < 0.01 Prior R vs. prior NR; t-tests

Fig. 3A shows the time in port during non-rewarded trials, as a function of the outcome of the previous trial (prior reward, or non-reward). A RM-ANOVA revealed a main effect of Prior Rewards (F(1, 38) = 27.144; P < 0.001) as prior rewards increased responding during subsequent non-rewarded trials. The analysis showed no main effect of Sex (F(1, 38) = 1.783; P = 0.190), but a marginal Sex x Prior Reward interaction (F(1, 38) = 4.015; P = 0.052), as males were slightly more influenced by prior rewards. The effect of trial history appears to be carried exclusively by the outcome of the previous trial (t-1) (no effect of consecutive prior rewards or non-rewards; Ps ≥ 0.147, Bonferroni t-tests; Fig. 3B).

We then analyzed another metric of conditioned responding: the response probability (i.e. probability subject enters the reward port during cue presentation). We reasoned that this metric might be more prone to detect brief lapses in discrimination performance (false alarms). Fig. 3C shows the probability of responding to a non-rewarded cue as a function of the outcome of the prior trial (prior reward or non-reward). A RM- ANOVA revealed a main effect of Prior Reward (F(1, 38) = 25.607; P < 0.001), and a significant Sex x Prior Reward interaction (F(1, 38) = 9.247; P = 0.004). In males, prior reward increased the probability of responding on subsequent non-rewarded trials (T(38) = 43.752; P < 0.001) while in females, trial history had no effect (T(38) = 1.631; P = 0.209). As a result, males and females differed in their probability of responding to non-rewarded trials when those trials were preceded by reward (T(38) = 4.508; P = 0.040). Here again we found no influence of the number of consecutive past reward or non-rewards (Ps ≥ 0.710, Fig. 3D).

Responding to rewarded cues remained high throughout the sessions and was not influenced by trial history (Ps ≥ 0.113); moreover no main effect or interaction with Sex was observed on these rewarded trials (Ps ≥ 0.127) (Fig. 3E-H).

These results indicate that male rats have difficulty inhibiting responding to non-rewarded (contextually irrelevant) cues following a recent reward. In contrast, female rats maintained high discrimination performance throughout the session and recent rewards had little to no effect on their discrimination accuracy. The mechanisms for the reward-induced decrease in discrimination accuracy (here observed primarily in males) are not fully understood. Rewards might temporarily lower the threshold for behavioral responding, resulting in stronger responding to weakly predictive cues [32]. Additionally, reward delivery might decrease discrimination accuracy by increasing the predictive value of the cue (and context) present during reward delivery. These direct cue→outcome (and context→outcome) associations compete with hierarchical associations for the control of behavior [33], potentially affecting responding on the next trial (e.g., a rewarded LIGHT: X+ trial might increase the value of cue X and promote responding during the following DARK: X- trial; the same logic could be applied to the influence of the context). Further investigation is required to determine the precise mechanism(s) by which reward disrupt discrimination performance in the context-dependent discrimination task.

Acute stress disrupts context-gated reward- seeking in male but not female rats

Stress affects several cognitive processes (e.g. selective attention, spatial navigation, working memory), often in a sex-dependent manner [25, 26, 34]. However the effect of stress on contextual modulation, and possible sex-differences in this effect, have not been formally tested. Therefore, we tested the effect of an acute stress (90-min restrain) on discrimination performance (expressed as discrimination ratio), in a subset of animals from all training groups (Ctx- dep. O1: n = 11 F + 15M; Simple discrimination: n= 6F + 6M; No discrimination: n= 8F +8M; Ctx-dep O1/O2: n= 7F + 7M) (Fig. 4). With the exception of the No Discrimination group, acute stress reduced discrimination accuracy in all tasks (main Stress effect: Ps ≤ 0.004). However, only in the Ctx-dep. O1 task did we observe a significant Stress x Sex interaction F(1, 24) = 5.665; P = 0.026) as only males showed reduced discrimination performance following acute stress (Stress vs. Baseline; Males: T(14) = 5.165; P < 0.001; Females: T(10) = 1.290; P = 0.209). No main effect or interaction with Sex was observed in any other task (Ps ≥ 0.178; note however the smaller sample size in all these other groups). Therefore, although acute stress disrupted discrimination performance in all tasks, females trained in Ctx-dep. O1 appear to be less sensitive to the disrupting effect of stress in this task. This is despite the fact that female rodents typically show greater hypothalamo-pituitary adrenal (HPA) activation in response to this type of stressor (i.e. greater stress activation) [36, 37].

Acute stress disrupts context-gated reward seeking in male but not female rats

(A-B) Time in port (% cue time) during the four different trial types (LIGHT: X / DARK: X / LIGHT: Y / DARK: Y) at baseline and after acute stress for male (A) and female (B) rats trained in each discrimination task. (C) Discrimination ratio at baseline and after acute stress, for male and female rats trained in each discrimination task. (D) Stress-induced disruption in discrimination performance (stress – baseline) for male and female rats trained in each discrimination task (horizontal bar = median difference score). Error bars indicate ± s.e.m; * P < 0.05; *** P < 0.001, t-tests.

Female rats show higher orbitofrontal engagement during context-gated reward predictions

The hierarchical gating of associative predictions has been linked to activity in the orbitofrontal cortex (OFC) [28–30]. Therefore, for a subset of animals (Ctx-dep. O1: n= 11 F + 9M; Simple discrimination: n= 7F + 9M; No discrimination: n= 7F +7M; Ctx-dep. O1/O2: n= 5F + 5M), we quantified c-Fos-immunoreactive (cfos+) cells in the OFC (medial, ventral and lateral subregions) following regular behavior testing (no-stress condition). We also quantified cfos+ in two control regions, M1J and S1J, where we did not expect task-specific activation. Compared to baseline (homecage controls), behavioral testing increased cfos+ expression in all regions of interest, regardless of task (Fig. 5; homecage c-Fos+ counts are shown but not included in the following statistical analyses). A RM-ANOVA revealed a main effect of Region F(4, 192) = 85.037; P ≤ 0.001) and a significant Task x Sex x Region interaction (F(12, 192) = 1.9; P = 0.036). Follow-up planned contrast analysis compared males and females within each task and brain region. In the Ctx-dep. O1 task, females showed increased c-fos+ expression compared to males, in all OFC subregions (T(14) ≥ 2.387; P ≤ 0.032). No sex differences were found in any other task (Ps ≥ 0.255; note however the smaller sample size of the Ctx-dep. O1/O2 group). Moreover, no sex differences were found in the control regions M1J and S1J, regardless of task (Ps ≥ 0.139).

Female rats show higher orbitofrontal engagement during context-gated reward seeking.

(A) Experimental design. Brains were collected 2h after the start of the behavioral session. Brain sections were stained for c-Fos (activity marker) and NeuN (general neuronal marker). (B) Behavioral performance (discrimination ratio) ratio during the final session, before tissue collection for male and female rats trained in each task. (C-G) c-Fos+ density in the primary jaw motor cortex (C), primary jaw somatosensory cortex (D), medial orbitofrontal cortex (E), ventral orbitofrontal cortex (F), and lateral orbitofrontal cortex (G) for males and females trained in each discrimination task. Error bars indicate ± s.e.m. Blue lines (close to floor level) indicate average c-fos+ density for animals who remained in their home cage prior to tissue collection (homecage controls; n=5). *: P < 0.05 Males vs. Females, t-tests.

Discussion

We investigated sex differences in contextual modulation of cue-evoked reward predictions and its neural correlates. Male and female rats were trained on a context-dependent discrimination task (Ctx-dep. O1) in which a contextual feature informs the validity of two reward cues (i.e. bidirectional occasion-setting). Overall, males acquired context-gated behavior faster than females. However, once established, contextual control over behavior was more robust in females; it was more stable within a session (less influence of prior reward) and more resistant to acute stress. This superior contextual control achieved by females was accompanied by elevated c-Fos expression in all OFC subregions (medial, ventral and lateral). Critically, no sex differences (behavioral or neurobiological) were observed among animals trained on simpler discrimination tasks that do not engage contextual gating (Simple Discrimination, No discrimination, Ctx- dep. O1/O2). These results add to the growing body of literature on sex differences in motivated behaviors and value-based decision-making [17, 24]. Most of these sex differences were observed in behaviors informed by binary associations between cues (or actions) and rewards. To our knowledge, this study is the first to demonstrate sex-differences in the context-informed hierarchical gating of associative predictions, a process that is largely independent from binary associations [3, 27].

For Ctx-dep. O1, the context (LIGHT or DARK chamber) is an occasion-setter, i.e. a stimulus that modulates the response to a target cue, presumably by gating the retrieval of the appropriate cue→outcome association. However, our task differs from classic occasion-setting preparations in several ways. Historically, Pavlovian modulation has been studied in preparations in which the modulating stimuli (the occasion-setter) exerts unidirectional modulation, either potentiating (A:X+ / X-) or reducing (A:X- / X+) the response to the target cue. In such preparations, the occasion- setter might also establish a direct association (excitatory or inhibitory) with the outcome [1, 38, 39], thereby entangling the contributions of hierarchical modulation and direct stimuli-evoked predictions. To circumvent this issue, occasion-setting preparations commonly introduce a temporal gap (5-30s) between the termination of a discrete occasion-setting stimulus and its target cue. This approach has generally been successful in limiting direct associations between the occasion-setter and the outcome, instead promoting hierarchical modulation processes [40, 41, but see 42]. This temporal gap however introduces a working- memory requirement (animals must remember if the target cue was preceded by the occasion-setter) which complicates results interpretations.

Here, we used a bidirectional occasion-setting preparation with a single reward outcome (Ctx- dep. O1) as an alternative approach to investigate contextual modulation [31]. In this paradigm, a contextual feature (e.g., LIGHT) positively modulates one target association (LIGHT: X+ / DARK: X-) and negatively modulates another association (LIGHT: Y- / DARK: Y+). Critically, there are no direct informative associations between contextual feature and the single reward outcome (rewards are equally frequent in the LIGHT or DARK contexts), so accurate discrimination necessitates hierarchical modulation [27]. Moreover, the contextual feature precedes and overlaps with the target cue, thereby eliminating working-memory requirements. Thus, sex differences observed in Ctx- dep. O1, strongly indicate sex differences in hierarchical (context-informed) associative modulations.

Alternatively, males and females might differ in terms of attentional resources [43]. Indeed, unlike Simple Discrimination or No discrimination tasks, Ctx-dep. O1 requires high multimodal attention (animals must simultaneously attend to the visual context and auditory cues). However, no sex differences were observed in context-dependent discrimination with two outcomes (Ctx-dep. O1/O2), a task that shares similar attentional requirements with Ctx-dep. O1, but does not necessarily engage hierarchical contextual modulation. Indeed, in Ctx-dep. O1/O2, both the visual context and the auditory target cue are relevant, however the use of differential outcomes presumably allows animals to solve the task via simple (nonhierarchical) summation processes, rats simply adding up the converging context-evoked and cue-evoked predictions. The possibility of a nonhierarchical solution to the discrimination problem might explain why almost all rats trained in the Ctx-dep. O1/O2 task learned the discrimination and at much faster rate compared to Ctx-dep. O1 (but see [31] for alternative explanation of the differential outcome effect). Together, these results indicate that the sex differences observed here are not attributable to simple associative, motivational, working-memory, or attentional processes, but are specific to the neurocomputational operations required for the hierarchical, contextual control of behavior. Note that there is no a priori reason to believe that the nature of these computations differ between sexes (no sex differences in the fundamental mechanisms of context- gated predictions). Rather male and female rats appear to differ slightly in the learning rate and magnitude/ robustness of these computations. Moreover it remains to be determined if the sex differences in context-gated reward seeking are specific to the sensory modalities employed here or if those sex differences extend to other sensory modalities.

Sex differences in context-dependent behaviors have previously been reported in fear conditioning preparations where females often display generalized fear responses that evade precise contextual control [44–47]. Elevated baseline anxiety and fear reactivity in females might contribute to this relative failure to regulate fear responses [48, 49, but see 50]. However, we show here that reduced contextual control in females (at least in the early stages of acquisition) is not limited to aversive behaviors but extends to appetitive, reward-seeking behaviors. This suggests a broader, domain-general, delay in the engagement of contextual, top-down regulation processes in female rodents. This is consistent with higher sign-tracking tendencies in female rats [51–55] —another form of appetitive behavior that reflects reduced top-down cognitive control (opposed to goal-tracking behavior) [56–58]. Note that the reduced top-down contextual control displayed by females in our study was limited to delayed acquisition. Provided sufficient training, females actually established more robust contextual control over cue-evoked reward seeking than males. The biological basis for the delayed acquisition of contextual control in females is unknown. Although estrous cycle had no effect on the expression of context- dependent behavior in female discriminators, fluctuating sex hormones (and resulting neurobiological changes) could have contributed to the slower acquisition of contextual control in females. Indeed, learning the hierarchical contextual rules for reward prediction is a complex and lengthy process that probably requires distributed plasticity across several brain regions including the striatum, amygdala, hippocampus and neocortex [3, 59]. Estrous cycle affects synaptic physiology and plasticity at all these sites via changes in dopamine reuptake [60, 61], neuronal excitability [62–66], spine density [67, 68], gene expression [69, 70], or neurogenesis [71]. This dynamic neurobiological environment might constitute a challenge for females attempting to solve the Ctx-dep. O1 task, and might contribute to their delayed acquisition of contextual control. Interestingly, this dynamic neurobiological environment might also contribute to the more stable performance ultimately achieved by females. Indeed, in deep learning models, the injection of random “noise” in (artificial) neural networks during training is commonly used to build redundancy and increase network performance and stability [72, 73]. This ‘robustness through perturbations’ approach often comes at the expense of learning speed, similar to what we observed in females. Alternatively, or in addition, androgens (male sex hormones) –known to contribute to the male advantage on certain spatial tasks- might have contributed to males’ faster acquisition of contextual control in this task [74, 75]. The role of these hormonal factors, as well as potential non-gonadal sex differences, in the trade-off between learning speed and performance stability remains to be confirmed experimentally.

Regardless of the precise mechanism, our results indicate that, compared to male rats, female discriminators ultimately achieved more stable contextual control over cued reward-seeking; their context-dependent discrimination performance was less affected by trial history (recent rewards) or by acute stress. This relative resilience of females trained in the Ctx-dep. O1 task is consistent with studies showing that females, while more sensitive to the arousing and anxiogenic effects of stress, are more resilient to the disrupting effects of stress on executive functions and memory [22, 26]. Sex differences in regional sensitivity to CRF and CRF receptor signaling might contribute to these sex differences in behavioral responses to stress [76].

The behavioral sex differences observed in this study were accompanied by neural sex differences. Compared to males, females displayed increased OFC activation during Ctx-dep. O1. Importantly, no behavioral or neural sex differences were observed for any other task. Together, these results strongly implicate the OFC in the sex-biased expression of contextual modulation of reward-seeking. This is consistent with recent studies demonstrating a critical role of the OFC in hierarchical control over cue-evoked behaviors [28–30]. More generally, the OFC is proposed to encode task structure including context-dependent contingencies [77–79]. Accordingly, the OFC appears critical for decision-making when optimal choices require knowledge of the broader task structure (i.e. when stimuli cannot be taken at face value) [80, 81]. Although causality remains to be established, our data suggest that the increased orbitofrontal engagement displayed by females might contribute to their resilient contextual regulation of behavior.

The root causes of the behavioral and neural sex differences in context-gated reward prediction are largely unknown and not addressed in this study. In principle, all the sex differences reported here (speed of acquisition, robustness of performance, and OFC activation) might be independent and engage different mechanisms. Alternatively, some of these effects might be causally related. Our result show that, for a comparable level of context-dependent discrimination, females show an increased engagement of the OFC compared to males (higher number of c-Fos immunoreactive neurons). This suggests a denser (less sparse) coding of context-dependent contingencies in females’ OFC. Denser coding is generally associated with slower discrimination learning but superior fault- tolerance (i.e. better resistance to noise or disturbance) [82], exactly what we observed in females, suggesting a potential unifying mechanism for the sex differences reported here.

The contextual regulation of behavior is at the heart of cognitive control [12, 83]. Deficits in this process are integral to the pathophysiology of numerous neuropsychiatric disorders [4]. Although extrapolation of animal data to humans requires caution, our findings of a superior OFC engagement by female rats during context-informed hierarchical control of reward seeking is consistent with human [f]MRI studies showing larger OFC grey matter volume in females [84] and superior OFC activation in females during the regulation of emotional responses [85, 86]. The behavioral and neural sex differences observed here might offer insight into the biological sex differences observed in certain disorders. For instance, the initial delay in establishing contextual control observed here in females might contribute to the rapid progression of SUD or PTSD observed in women following initial drug use or trauma [14, 15, 87, 88]. Conversely, the more stable contextual control ultimately achieved by females after extended training observed here might contribute to the better prognosis for women following behavioral and cognitive treatments for these disorders [88–90], and might be due, in part, to increased OFC engagement.

Methods

Subjects

Subjects were male and female Long Evans rats (Charles Rivers), 8-10 weeks at arrival. Rats were housed in same-sex pairs in a temperature-controlled vivarium (22 °C) with a 12-hr light/dark cycle; behavioral experiments were conducted during the light cycle. Rats were given free access to water in their home cage for the duration of the experiments. After a week of acclimation to the colony room, rats were mildly food-restricted to maintain ∼95% of age-matched free- feeding weights. All experimental procedures were conducted in accordance with UCSB Institutional Animal Care and Use Committees and the US National Institute of Health guidelines.

Apparatus

Behavioral training was conducted in 12 identical conditioning chambers enclosed in individual sound- and light-attenuating cubicles (Med Associates, St. Albans, VT). Six chambers were dedicated to male subjects and six chambers were dedicated to female subjects; animals of both sexes were trained simultaneously. A recessed reward delivery port was located at the center of the front panel and was connected to a pellet dispenser and a syringe pump for liquid reward delivery (syringe pump located outside of the sound-attenuating cubicle). Subjects’ presence in the reward port was detected by the interruption of an infrared beam. Two ceiling-facing chamber lights located on the front and back internal walls of the cubicle provided diffuse chamber illumination and were used to manipulate the background visual context. Auditory stimuli were delivered via a white noise speaker and a clicker (76dB each), located in the back and front panel, respectively. A fan mounted on the cubicle provided ventilation and low background noise. A computer equipped with Med-PC software (Med Associates) controlled all experimental programs and data collection.

Rewards

Food rewards consisted of a 45mg banana-flavored food pellet (BioServe Dustless Precision Pellets, F0059 formula) or a 0.17mL bolus of sweetened chocolate milk (2 parts chocolate Nesquik® + 1 part water + 5% sucrose w/v) delivered over 3s. These rewards are highly discriminable but equicaloric and were equally preferred by rats in preliminary studies. The identity of the rewarding outcomes was counterbalanced within behavioral tasks and within sexes.

Discrimination Training

Rats were randomly assigned to one of four Pavlovian discrimination conditions: Context dependent discrimination with single outcome (Ctx-dep. O1, n = 30F + 32M), Context dependent discrimination with dual outcome (Ctx-dep. O1/O2, n = 14F + 14M), Simple discrimination (Simple, n = 12F + 22M), or No Discrimination (NoDiscr., n = 8F + 8M). Animals were individually placed in the conditioning chambers for 80-min training sessions 5-6 days/week. During each session, the visual background –the local context– alternated every two minutes between flashing houselights (0.25s on, 0.25s off; LIGHT context) or darkness (DARK context). Within each context phase, two 10-s auditory cues, X or Y (white noise or clicker, counterbalanced), were presented individually, in two separate trials, and were potentially followed by reward. The presentation of each cue within a context phase was selected semi-randomly, so that the identity of the first cue had no incidence on the identity of the second cue (i.e. the same cue could repeat twice within a context phase, so the outcome of the first trial did not inform the outcome of the second trial). Each session consisted of 40 context cycles and 80 cue presentations (i.e. 80 trials) with an average intertrial of 50s ± 20s (rectangular distribution). The reward contingencies in each task are outlined in Figure 1B. During the acquisition period, unless specified otherwise (cf. Ctx- dep. O1), each behavioral training session consisted of an equal proportion of each trial type (20* LIGHT: X; 20* DARK: X; 20* LIGHT: Y; 20* DARK: Y).

Context dependent discrimination with single outcome (Ctx-dep. O1)

In this task, the reward-predictive validity of each cue was informed by the context, so that cue X was rewarded only in LIGHT context and cue Y was rewarded only in the DARK context. All rewarded trials resulted in the same reward outcome (O1). Considered individually, both cues and both contexts have equal probabilities of resulting in the same outcome, therefore accurate performance in this task cannot be understood in terms of summed associative strength between cues and contexts. Instead, accurate performance in this task critically relies on contextual modulation (i.e. occasion-setting). During the first 35 training sessions, rats experienced an equal number of rewarded and unrewarded trials (20 trials for each context/cue combination per session). On the 36th session and for the remainder of the experiment, the relative proportion of rewarded trials was reduced (10* LIGHT: X+; 30* DARK: X-; 30* LIGHT: Y-; 10* DARK: Y+) in an effort to promote discrimination. After 72 training sessions, animals that failed to reach a discrimination criterion were excluded from the rest of the study (discrimination criterion: time in port during rewarded cue ≥ 1.5* time in port during nonrewarded cue, for both cues, for a minimum of 4 out of 5 consecutive sessions).

Simple Discrimination

Cue X was always rewarded and cue Y was never rewarded, regardless of the context. Presumably, this training protocol promotes accurate reward prediction via simple associative learning (context-independent predictions).

No Discrimination

Both cues X and Y were probabilistically rewarded in both context (the background context provides no disambiguating information). During the acquisition period, the probability of reinforcement was 50% for each cue/ context combination. Presumably, this training protocol promotes reward prediction via simple associative learning (context-independent predictions) although the reward cues retain some ambiguity.

Context dependent discrimination with dual outcome (Ctx-dep. O1/O2)

This task followed similar contingencies as Ctx-dep. O1, with the exception that the two rewarded trial types (LIGHT: X and DARK: Y) resulted in two different food outcomes (O1 and O2, respectively). This has profound implications for the learning processes engaged in the task. The delivery of different outcomes in different contexts allows for distinct context-outcome associations (LIGHT→O1 and DARK→O2), as well as distinct cue-outcome associations (X→O1 and Y→O2). These different associative structures could potentially facilitate discrimination performance via simple summation process.

Assessment of Estrous Cycle

Estrous cycle phases were monitored via vaginal cytology for a subset of females (n = 10) over 15 consecutive days (sessions 66-80). Immediately after the behavioral session, the tip of a saline-moistened cotton swab was inserted into the vagina and rotated to dislodge cells from the vaginal wall. Swabs were immediately rolled onto a glass slide, and the samples were preserved with a spray fixative (M-FixTM, MilliporeSigma) without allowing the cells to dry. Slides were then stained with a modified Papanicolau staining procedure as follows: 50% ethyl alcohol, 3 min; tap water, 10 dips (×2); Gill’s hematoxylin 1, 6 min; tap water, 10 dips (×2); Scott’s water, 4 min; tap water, 10 dips (×2); 95% ethyl alcohol, 10 dips (×2); modified orange-greenish 6 (OG-6), 1 min; 95% ethyl alcohol, 10 dips; 95% ethyl alcohol 8 dips; 95% ethyl alcohol 6 dips; modified eosin azure 36 (EA-36), 20 min; 95% ethyl alcohol, 40 dips; 95% ethyl alcohol, 30 dips; 95% ethyl alcohol, 20 dips; 100% ethyl alcohol, 10 dips (×2); xylene, 10 dips (×2); coverslip immediately. Samples were observed with light microscopy (4x). Four criteria were used to determine the specific estrus cycle: 1. cell density, 2. percentage of nucleated epithelial cells, 3. cornified epithelial cells, and 4. Percentage of leukocytes [88–90]. Each sample was independently rated by two trained observers (S.P. and A.M.), blind to the behavioral data. Inter-rater agreement was high (88.2%). Only behavioral sessions for which both raters agreed on the estrous phase were selected for analysis. As a result, 2-4 behavioral sessions were considered for each subject at each phase of the estrous cycle. Behavioral data on these sessions were averaged to express a single data point for each rat at each stage of the estrous cycle. Females that did not show a regular cycle were excluded from the analysis (n = 2).

Acute Restraint Stress

Before testing the effect of acute stress on discrimination performance, the relative proportion of rewarded trials was adjusted in all groups to match the proportion of rewarded trials in the context-dependent discrimination with a single outcome (Ctx-dep. O1). As a result, rewarded cues became less frequent and represented only 1/4 of the trials. For all groups, these rewarded trials were equally distributed across both contexts. After 5-6 sessions of adjustment to these new conditions, we tested the effect of an acute stress on discrimination performance in a subset of animals from all training groups (Ctx-dep. O1: M n=15, F n=11; Simple discrimination: M n=6, F n=6; No discrimination: M n=8, F n=8; Ctx-dep. O1/O2: M n=7, F n=7). Acute stress was induced by restraining rats in standard Plexiglas rat restraint tubes (Plas Labs, IncTM) for 90 min, in a novel brightly lit room. To account for the size differences between male and females, two different sized restrainers were used (65 and 50mm internal diameter for male and female restrainers respectively). After placing the rats in the tube, the length of the restrainer was adjusted to immobilize the rat without causing pain or interfering with animals’ breathing. After 90 min of restraint, the rats were returned to their homecage and left undisturbed for 5 min before being placed in the conditioning chambers for a normal behavioral session. This restraint procedure is a well- established stressor in rodents, known to produce robust behavioral and physiological effects [35–37].

cFos Immunofluorescence

Following stress test, animal were retrained for 5-6 sessions, under standard conditions (no stress; equal number of rewards for all training groups; rewarded trials = 1/4 of all trials). After their last behavioral session, animals remained in the conditioning chambers for 40 min (120 min total); they were then deeply anesthetized with a lethal dose of Euthasol and perfused transcardially with ice-cold 1x PBS followed by ice-cold 4% PFA in PB. Brains were extracted, post-fixed (4% PFA overnight) and cryoprotected (30% sucrose + 0.1% sodium azide, 1x PBS) until processed for immunostaining. Brains were frozen and sectioned into 40μm with a sliding cryostat (Leica CM1860). Free floating coronal sections were washed for 10 min in PBS, followed by a 2-h incubation in a blocking solution (5% NGS, 0.2% Triton X-100 in 1x PBS) at room temperature. Sections were then incubated with rabbit anti-cFos (1:1000, Cell Signaling, CAT#: 2250) and mouse anti-NeuN (neuron-specific protein; 1:2000 or 1:3000, Novus, CAT#: NBP1-92693) primary antibodies, in the blocking solution, for 20 h a 4°C. Following primary antibody incubation, sections were washed in PBS 3 times for 10 min and then incubated with anti-rabbit Alexa Fluor® 647-conjugated secondary antibody (far-red) (1:250, Jackson ImmunoResearch, CAT#: 111-605-144) and anti-mouse Alexa Fluor® 488-conjugated secondary antibody (green) (1:500, Jackson ImmunoResearch, CAT#: 115-545-003) for 90 min at room temperature. Sections were then washed in PBS three times for 10 min, mounted on slides and immediately coverslipped with a hardset mounting medium (Vector laboratories CAT#: H-1700). Images were acquired at a 10X magnification, using a Keyence BZ-X fluorescence microscope.

Brain regions of interest for c-Fos immunoreactive neurons quantification included the medial, ventral, and lateral orbitofrontal cortex (mOFC, vOFC, and lOFC). Additionally, two control regions were included, for which we did not expect task-specific activation: the jaw primary motor (M1J) and somatosensory cortex (S1J). These regions of interests were manually isolated from the original images, using only the green / NeuN channel (the far-red / c-fos channel was temporarily disabled for blinding purposes). A custom ImageJ script (National Institute of Health) was then used to count c-Fos positive neurons in the selected regions and normalize this count to the surface area of the selected region (c-Fos positive neurons / mm2). For each brain region, c-Fos positive neurons were counted bilaterally from 2-4 sections per animal and a mean count was computed for each animal.

Statistical analysis

Data were collected over three replication cohorts, each composed of approximately equal number of male and female rats. Cue-evoked reward seeking was quantified primarily as the percentage of time a rat spent in the food cup during the last 5 seconds of cue presentation, when anticipatory food cup behavior is more reliably observed and less contaminated by orienting behaviors [91–93]. The percentage of time in port during the 5s epoch prior cue onset is shown but not analyzed. Discrimination accuracy was expressed as a discrimination ratio, defined as the time in port during rewarded trials divided by the total time in port during all trials. A ratio of 0.5 indicates equal responding during rewarded and non-rewarded trials (i.e. no discrimination) whereas a ratio of 1 indicates responding exclusively during rewarded trials (i.e. perfect discrimination). Time in port and discrimination ratio during acquisition were aggregated in 4-session bins. These longitudinal repeated measures were analyzed with Linear Mixed-Models (LMM) with a Restricted Maximum Likelihood (REML) estimation method. All covariance structures were explored and the best fitting model was determined by the lower Akaike information criterion score [94, 95]. The selection of a covariance structure influenced the degrees of freedom of the model, possibly resulting in non-integer values. The remaining of the analyses consisted generally of mixed-models repeated measures (RM) ANOVAs with Sex and/or Task as a between-subject factor, and Trial history, Stress, or Brain region as within-subject factors. The Geisser-Greenhouse correction was applied when data lacked sphericity, possibly resulting in non-integer degrees of freedom. Post-hoc or planned contrast pairwise comparisons were carried with Bonferroni- corrected t-tests. When appropriate, nonparametric tests were conducted and consisted of the Kruskal- Wallis H test followed by post-hoc Bonferroni-corrected Dunn’s pairwise comparisons. All tests were two-tailed, significance was assessed against a type I error rate of 0.05. Statistical analyses were conducted with SPSS Statistics package (version 28.0.0.0; IBM SPSS). Graphs were generated using GraphPad Prism 9.

Supporting information